---
title: "Box-Whisker Plots"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Box-Whisker Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 300,
  fig.show = "hold",
  fig.align = "center",
  fig.width = 7,
  out.width = "80%",
  comment = "#>"
)
```

# 1. Introduction

The following vignette aims at documenting and illustrating workflows for producing box-and-whisker plots using the `ospsuite.plots`-library.

The function for plotting box-whiskers is: `plotBoxWhisker`. 
Basic documentation of the function can be found using: `?plotBoxWhisker`.
The output of the function is a `ggplot` object.

## 1.1 Setup

For this vignette we will use the libraries {ospsuite.plots} and {tidyr}.
We use the default settings of {ospsuite.plots} (see vignette("ospsuite.plots", package = "ospsuite.plots")) but adjust the legend position and the alignment of the caption.


```{r setup,warning=FALSE , message=FALSE}
library(ospsuite.plots)
library(tidyr)

# setDefaults
oldDefaults <- ospsuite.plots::setDefaults()


# Adjust default theme plot for prettier  plots
theme_update(legend.position = "top")
theme_update(legend.direction = "horizontal")
theme_update(plot.caption = element_text(hjust = 1))
```



## 1.2 Example data

This vignette uses a data set provided by the package:


```{r load-data, results='asis'}
pkRatioData <- exampleDataCovariates %>%
  dplyr::filter(SetID == "DataSet1") %>%
  dplyr::select(-c("SetID", "gsd", "AgeBin")) %>%
  dplyr::mutate(Agegroup = cut(Age,
    breaks = c(0, 6, 12, 18, 60),
    include.lowest = TRUE,
    labels = c("infants", "school children", "adolescents", "adults")
  ))


knitr::kable(
  head(pkRatioData),
  digits = 3,
  caption =
    "first rows of example data pkRatioData"
)
```


Metadata is a list, which contains dimension and unit information for dataset columns.
If available axis labels are set by this information.


```{r meta-data, results='asis'}
metaData <- attr(exampleDataCovariates, "metaData")


knitr::kable(metaData2DataFrame(metaData), digits = 2, caption = "list of meta data")
```


# 2. Examples


## 2.1 Examples for aggregation of categorical data

### 2.1.1 Minimal example

Age (mapped to y) is aggregated.

```{r minimal-example}
plotBoxWhisker(
  data = pkRatioData,
  mapping = aes(y = Age),
  metaData = metaData
)
```

### 2.1.2 Stratification on x-axis

Age (mapped to y) is aggregated for different countries (mapped to x)

```{r example-x-stratification}
plotBoxWhisker(
  mapping = aes(
    x = Country,
    y = Age
  ),
  data = pkRatioData,
  metaData = metaData
)
```

### 2.1.3 Stratification by color

Age (mapped to y) is aggregated for different countries (mapped to fill)

```{r minimal-example-color}
plotBoxWhisker(
  mapping = aes(
    fill = Country,
    y = Age
  ),
  data = pkRatioData,
  metaData = metaData
)
```

### 2.1.4 Stratification by color and on x-axis

Age (mapped to `y`) is aggregated for different countries  (mapped to `x`) and Sex (mapped to `groupby`). `groupby` is an additional aesthetics of {ospsuite.plots} which works together with the variable `groupAesthetics`. For the function `plotBoxwhisker()`  `groupAesthetics` is not settable and fixed to `fill`.

```{r minimal-example-color-x}
plotBoxWhisker(
  mapping = aes(
    x = Country,
    y = Age,
    groupby = Sex
  ),
  data = pkRatioData,
  metaData = metaData,
)
```

### 2.1.4 Stratification by column combination

Below `groupby`was mapped to a combination of the columns "Sex" and "Country":

```{r minimal-example-interaction}
plotBoxWhisker(
  mapping = aes(
    y = Age,
    groupby = interaction(Country, Sex, sep = "-")
  ),
  data = pkRatioData,
  metaData = metaData
) +
  theme(legend.title = element_blank())
```


### 2.1.6 Omit data points flagged as Missing Dependent Variable (mdv)

If some of the data should be omitted, we could do this by mapping a `logcial` column to the aesthetic `mdv`. Below we excluded data from Germany: 

```{r minimal-example-mdv}
plotBoxWhisker(
  mapping = aes(
    x = Country,
    y = Age,
    mdv = Country == "Germany",
    groupby = Sex
  ),
  data = pkRatioData,
  metaData = metaData
)
```


## 2.2 Examples for box-whisker vs  numeric data 

### 2.2.1 Numeric data as factor

In the next example we added a numeric column as "mean age" of the age group to the data set.
This column is mapped *as factor* to `x`. The values are now displayed like categorical values equidistant:


```{r numeric-example-factor}
pkRatioData <- pkRatioData %>%
  dplyr::group_by(Agegroup) %>%
  dplyr::mutate(meanAge = round(mean(Age), 2))

metaData[["meanAge"]] <- metaData[["Age"]]
metaData <- metaData[intersect(names(pkRatioData), names(metaData))]

plotBoxWhisker(
  data = pkRatioData,
  mapping = aes(
    x = as.factor(meanAge),
    y = Ratio,
    fill = Agegroup
  ),
  metaData = metaData
)
```

### 2.2.2 Numeric data as distinct numeric values

If the column mapped to x is numeric and not a factor, the x-position of the boxes correspond to the numeric value:


```{r numeric-example-distinct-values}
plotBoxWhisker(
  mapping = aes(
    x = meanAge,
    y = Ratio,
    fill = Agegroup
  ),
  data = pkRatioData,
  metaData = metaData
)
```

### 2.2.3 Map continuous column with binning 

By mapping x with a function, it is also possible to aggregate data of a continuous column to bin.  
Below we use the function `cut` to aggregate the data in 4 distinct bin (see ?cut).  

Attention: for `cut(Age)` no `metaData` exists. So we have to set the x label manually


```{r numeric-example-bin4}
plotBoxWhisker(
  mapping = aes(
    x = cut(Age, 4),
    y = Ratio
  ),
  data = pkRatioData,
  metaData = metaData
) +
  labs(x = "Age range [years]")
```

We can use any functions which converts a continuous vector to a factor.
Below we define our own cut function simply as a wrapper around cut, with some fixed arguments.  
This function is then mapped to `x` and `groupby`.

```{r numeric-example-binfunction}
myCutfun <- function(x) {
  cut(
    x = x,
    breaks = c(0, 6, 12, 18, 60),
    include.lowest = TRUE,
    labels = c("infants", "school children", "adolescents", "adults")
  )
}


plotBoxWhisker(
  mapping = aes(
    x = myCutfun(Age),
    y = Ratio,
    groupby = myCutfun(Age)
  ),
  data = pkRatioData,
  metaData = metaData
) +
  theme(legend.title = element_blank()) +
  labs(x = "")
```



## 2.3. Aggregation function

By default the data is aggregated by percentiles defined by the default option `ospsuite.plots.Percentiles`.
The percentiles can be customized for a specific plot by the input variable `percentiles` or for all plots generated by `plotBoxwhisker()` by changing the default options (`setOspsuite.plots.option(optionKey = OptionKeys$Percentiles, value =  c(0.05, 0.25, 0.5, 0.75, 0.95))`).

It is also possible to use a customized function via the input variable `statFun`. If `statFun` is not `NULL` it will overwrite the percentiles.


**Important**: If you override the defaults this way, please make sure to specify this in the plot annotations as you are basically redefining a box plot and the reader might not be aware of this and will misinterpret the plot.

## 2.3.1 Example for customized percentiles

In the example below we do not show the whiskers, by setting whisker percentiles to box percentiles

```{r customized-percentiles}
# B  No whiskers
plotBoxWhisker(
  mapping = aes(
    x = Country,
    y = Age
  ),
  data = pkRatioData,
  metaData = metaData,
  percentiles = c(0.25, 0.25, 0.5, 0.75, 0.75)
)
```

## 2.3.2 Example for customized aggregation function

To customize the aggregation provide a function which has as input the vector to aggregate and as output a named list with entries `ymin`,`lower`,`middle`,`upper`,`ymax`
In the example below  the function `myStatFun` was provided, which uses mean and standard deviation to aggregate:


```{r aggregation-functions}
myStatFun <- function(y) {
  r <- list(
    ymin = mean(y) - 1.96 * stats::sd(y),
    lower = mean(y) - stats::sd(y),
    middle = mean(y),
    upper = mean(y) + stats::sd(y),
    ymax = mean(y) + 1.96 * stats::sd(y)
  )
  return(r)
}

plotBoxWhisker(
  mapping = aes(
    x = Country,
    y = Age
  ),
  data = pkRatioData,
  metaData = metaData,
  statFun = myStatFun
) +
  labs(
    caption = "mean for the middle line,
       mean +/- standard deviation for the box edges
       and mean +/- 1.96 standard deviation for the whiskers"
  )
```

## 2.3.3 Tables corresponding to plot

As there already exist a lot off possibilities to aggregate data in R, this package does not provide an extra function to return the aggregated data in a tabular form.  

But the `plotBoxWhisker()` adds the aggregation function to the plot object. So the aggregation can be done easily e.g. with package {data.table} using the same function as used for the plot.

To be sure to use the same data use `plotObject$data`, then applied data filters by aesthetic `mdv` will be accounted for.

```{r aggregationTable}
# generate plot object
plotObject <- plotBoxWhisker(
  mapping = aes(
    x = Country,
    y = Age,
    fill = Sex
  ),
  data = pkRatioData,
  metaData = metaData
)

plot(plotObject)

# convert data to data.table and use  statFun saved in plotObject for aggregation
# make sure to add all relevant aesthetics to "by" columns
dt <- plotObject$data %>%
  data.table::setDT() %>%
  .[, as.list(plotObject$statFun(Age)),
    by = c("Country", "Sex")
  ]

knitr::kable(dt)
```




## 2.4. Show Outlier


### 2.4.1 Outlier with default settings


Outliers are displayed if variable `outliers` is TRUE. 
Default outliers are flagged when outside the range from "25th" percentile - 1.5 x IQR to "75th" percentiles + 1.5 x IQR.


```{r example-outliers}
plotBoxWhisker(
  mapping = aes(
    x = Sex,
    y = Ratio,
    fill = Country
  ),
  data = pkRatioData,
  metaData = metaData,
  outliers = TRUE
) +
  labs(
    tag = "A",
    caption = "Default settings: Whisker indicate 90% range (5th - 95th percentile)
       and outlier indicate all measurements outside
       25th percentiles - 1.5 x IQR to 75th percentiles + 1.5 x IQR"
  )
```

### 2.4.2 Outlier with default settings



In the following example the aggregation is customized to set the whiskers to 10% and 90% via the input variable percentiles and the outlier range is customized to show all points outside the whiskers.  
For that we define a function `myStatFunOutlier` which has as input the vector of the values, and as output the vector of values outside the outlier range.  
If no value is outside range NA is returned.


```{r example-outliers-custom}
myStatFunOutlier <-
  function(y) {
    q <-
      stats::quantile(y,
        probs = c(0.1, 0.9),
        names = FALSE,
        na.rm = TRUE
      )
    yOutsideRange <- subset(y, y < q[1] |
      y > q[2])
    if (length(yOutsideRange) < 1) {
      return(as.double(NA))
    } else {
      return(yOutsideRange)
    }
  }


plotBoxWhisker(
  data = pkRatioData,
  metaData = metaData,
  mapping = aes(
    x = Sex,
    y = Ratio,
    fill = Country
  ),
  outliers = TRUE,
  percentiles = c(0.1, 0.25, 0.5, 0.75, 0.9),
  statFunOutlier = myStatFunOutlier
) +
  labs(
    caption = "Whisker indicate 80% range (10th - 90th percentile)
       and outlier indicate all measurements outside whiskers"
  )
```

## 2.4.3 Retrieve outlier as table

The `plotBoxWhisker()` adds the outlier function to the plot object. So outliers can be easily retrieved e.g. with package {data.table}


```{r outlierTable}
# generate plotObject
plotObject <- plotBoxWhisker(
  data = pkRatioData,
  metaData = metaData,
  mapping = aes(
    x = Sex,
    y = Ratio,
    fill = Country
  ),
  outliers = TRUE,
  percentiles = c(0.1, 0.25, 0.5, 0.75, 0.9),
  statFunOutlier = myStatFunOutlier
)

# convert data to data.table and use  statFun saved in plotObject for aggregation
# make sure to add all relevant aesthetics to "by" columns
dt <- plotObject$data %>%
  data.table::setDT() %>%
  .[, .(outliers = plotObject$statFunOutlier(Ratio)),
    by = c("Country", "Sex")
  ]

knitr::kable(dt)
```




# 3. Plot configuration of boxplots

Below you see an example of the same plot  

* A default layout
* B customized layout

  * the variable `geomBoxplotAttributes`  has now an entry for color 'orange'.
  * the variable `geomPointAttributes`  has now an entry for size = 4, color 'orange' and shape = 'diamond'.
  * fill was added to the mapping so the outliers points are also colored corresponding to 'Sex'
  * with `scale_fill_manual`  the colors for the 'Sex' are defined
       



```{r boxplot-update-configuration}
# A default layout
plotBoxWhisker(
  data = pkRatioData,
  metaData = metaData,
  mapping = aes(
    x = Country,
    y = Age,
    fill = Sex
  ),
  outliers = TRUE,
  percentiles = c(0.1, 0.25, 0.5, 0.75, 0.9),
  statFunOutlier = myStatFunOutlier
) +
  labs(tag = "A")


# B customized layout
plotBoxWhisker(
  data = pkRatioData,
  metaData = metaData,
  mapping = aes(
    x = Country,
    y = Age,
    fill = Sex
  ),
  outliers = TRUE,
  percentiles = c(0.1, 0.25, 0.5, 0.75, 0.9),
  statFunOutlier = myStatFunOutlier,
  geomBoxplotAttributes = list(
    color = "orange",
    position = position_dodge(width = 1)
  ),
  geomPointAttributes = list(
    position = position_dodge(width = 1),
    size = 4,
    color = "orange",
    shape = "diamond filled"
  )
) +
  scale_fill_manual(values = c(
    Female = "pink",
    Male = "dodgerblue"
  )) +
  labs(tag = "B")
```

```{r cleanup,echo=FALSE}
resetDefaults(oldDefaults)
```

