---
title: "Goodness of fit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Goodness of fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 300,
  fig.show = "hold",
  fig.align = "center",
  fig.width = 7,
  out.width = "80%",
  comment = "#>"
)
```



# 1. Introduction

The `ospsuite.plots`-library provides functions to compare predicted and observed data and resulting residuals: 

* `plotPredVsObs()` (see section 2)  plot predicted values v corresponding observed values.
* `plotResVsCov()` (see section 3) plot residuals as point vs a covariate (e.g Time, observed values or Age)
* `plotRatioVsCov()` (see section 4) plot ratios  as point vs a covariate (e.g Time, observed values or Age)
* `plotQQ()` (see section 5) (Quantile Quantile plot)

In these functions the aesthetics `observed` and `predicted` can be used.
These aesthetics are available also in other functions of the `ospsuite.plots`-library:  

* `plotBoxwhisker()` (see section 6.1)  aggregate residuals and display aggregated values vs covariate
* `plotHistogram()` (see section 6.2)   generates histograms of residuals


The functions `plotPredVsObs()`, `plotResVsCov()`and `plotRatioVsCov` are all base on `plotXVsY()`. 
Mostly the default of the inputs are changed or the inputs are fixed. So please also use `?plotXvsY`to get more Details.

For DDI Comparison  the functions `plotPredVsObs()` and `plotResVsObs()` can be overlayed by the lines indicating the limits of the Guest Criteria  (<https://dmd.aspetjournals.org/content/39/2/170>) (see section 7)

## 1.1 Setup

For this vignette we will use the libraries {ospsuite.plots} and {tidyr}.
We use the default settings of {ospsuite.plots} (see vignette("ospsuite.plots", package = "ospsuite.plots"))


```{r setup, warning=FALSE,message=FALSE}
library(ospsuite.plots)
library(tidyr)

oldDefaults <- setDefaults()
```

## 1.2 Example data

This vignette uses two random generated example data sets provided by the package.

### 1.2.1 Dataset with predicted and observed data

```{r load-data-1, results='asis'}
data <- exampleDataCovariates %>%
  dplyr::filter(SetID == "DataSet2") %>%
  dplyr::select(c("ID", "Age", "Obs", "gsd", "Pred", "Sex"))

knitr::kable(
  head(data),
  digits = 3,
  caption =
    "first rows of example data "
)


metaData <- attr(exampleDataCovariates, "metaData")
metaData <- metaData[intersect(names(data), names(metaData))]

knitr::kable(metaData2DataFrame(metaData), digits = 2, caption = "list of meta data")
```

### 1.2.2 Dataset for examples with DDI prediction


```{r load-data-2, results='asis'}
# Load example
DDIdata <- exampleDataCovariates %>%
  dplyr::filter(SetID == "DataSet3") %>%
  dplyr::select(c("ID", "Obs", "Pred")) %>% 
  dplyr::mutate(Study = paste('Study',ID))

DDIdata$Study = factor(DDIdata$Study,levels = unique(DDIdata$Study))

knitr::kable(head(DDIdata), digits = 2, caption = "first rows of dataset used for DDI example")

DDImetaData <- list(
  Obs = list(
    dimension = "DDI AUC Ratio",
    unit = ""
  ),
  Pred = list(
    dimension = "DDI AUC Ratio",
    unit = ""
  )
)


knitr::kable(metaData2DataFrame(DDImetaData), digits = 2, caption = "list of meta data")
```

## 1.2.3 Dataset for Ratio comparison


```{r load-data-3, results='asis'}
# Load example
pkRatioData <- exampleDataCovariates %>%
  dplyr::filter(SetID == "DataSet1") %>%
  dplyr::select(-c("SetID")) %>%
  dplyr::mutate(gsd = 1.1)


pkRatiometaData <- attr(exampleDataCovariates, "metaData")
pkRatiometaData <- metaData[intersect(names(pkRatioData), names(pkRatiometaData))]

knitr::kable(head(pkRatioData), digits = 3)

knitr::kable(metaData2DataFrame(pkRatiometaData), digits = 3)
```


# 2. Predicted vs Observed (`plotPredVsObs()`)

## 2.1 Basic examples 

### 2.1.1 Default settings 

Basic example using default settings. 
Predicted and observed data are mapped with `predicted` and `observed`.  
Aesthetic `groupby` can be used to group observations.


```{r basic-examples}
plotPredVsObs(
  data = data,
  mapping = aes(
    observed = Obs,
    predicted = Pred,
    groupby = Sex
  ),
  metaData = metaData
)
```


### 2.1.2 Basic example linear scale

 The scale for the x and y axis is set to linear. It is not intended to use different scales for x and y axis. Therefore only one variable `xyscale` exists for both axis.  

Predicted and observed data are mapped with `x` and `y`. 
 

```{r basic-examples-linear}
plotPredVsObs(
  data = data,
  mapping = aes(
    x = Obs,
    y = Pred,
    groupby = Sex
  ),
  metaData = metaData,
  xyscale = "linear"
)
```

### 2.1.3 Error bars for observed data

For the observed data error bars are plotted. In the example this is done by mapping `error_relative.` Error bars could also be produced by mapping `error` to a column with an additive error or mapping `xmin` and `xmax` explicitly.



```{r basic-examples-errorbar}
plotPredVsObs(
  data = data,
  mapping = aes(
    x = Obs,
    y = Pred,
    error_relative = gsd,
    groupby = Sex
  ),
  metaData = metaData
)
```


### 2.1.4 Error bars for observed and predicted data

To plot the prediction error also `ymin` and `ymax`  have to be mapped explicitly

```{r basic-exampleserror-y}
plotPredVsObs(
  data = data,
  mapping = aes(
    x = Obs,
    y = Pred,
    error_relative = gsd,
    ymin = Pred * 0.9,
    ymax = Pred * 1.1,
    groupby = Sex
  ),
  metaData = metaData
)
```

### 2.1.5 Example with LLOQ

Below a data set is created where the LLOQ is set to the 0.1 quantile of the observed data. All values below are set to a LLOQ/2.
By mapping lloq this data are displayed with a lighter alpha, And a horizontal line for the lloq is added

```{r basic-examples-lloq}
LLOQ <- signif(quantile(data$Obs, probs = 0.1), 1)

dataLLOQ <- data %>%
  dplyr::mutate(lloq = LLOQ) %>%
  dplyr::mutate(Obs = ifelse(Obs <= lloq, lloq / 2, Obs))

plotPredVsObs(
  data = dataLLOQ,
  mapping = aes(
    x = Obs,
    y = Pred,
    lloq = lloq,
    groupby = Sex
  ),
  metaData = metaData
)
```


## 2.2 Adjust comparison lines  

### 2.2.1 Adjust fold distance


`plotPredVsObs()` adds lines to indicates fold distances.
The lines are defined by variable `comparisonLineVector`, which is a named list with default values
`list(identity =1, '1.5 fold' = c(1.5, 1/1.5),'2 fold' = c(2, 1/2)`.

A fold distance list can be generated by the helper function `?getFoldDistanceList()`.  


Below the 1.2 and 1.5 distances are displayed:

```{r examples-changeOfFold}
plotPredVsObs(
  data = data,
  mapping = aes(
    x = Obs,
    y = Pred,
    groupby = Sex
  ),
  metaData = metaData,
  comparisonLineVector = getFoldDistanceList(c(1.2, 1.5))
)
```

### 2.2.2 Adjust display of lines

The names of the list are displayed in the legend.  

If the list is unnamed all lines are displayed with the same `linetype` and they are not included in the legend. 
The linetype then used is settable by the variable `geomComparisonLineAttributes`

If the variable `comparisonLineVector` is NULL no lines will be displayed.  


```{r examples-adjustFOldlegend}
plotPredVsObs(
  data = data,
  mapping = aes(
    x = Obs,
    y = Pred,
    groupby = Sex
  ),
  metaData = metaData,
  comparisonLineVector = unname(getFoldDistanceList(c(1.2, 1.5))),
  geomComparisonLineAttributes = list(linetype = "dotted")
)
```



## 2.3 Adding a regression line


To add a regression line set the input variable `addRegression` to `TRUE` (A).  
For regression lines the package {ggpubr} has a nice function `stat_regline_equation` to add statistics of the regression as labels (B).
To use other function e.g. local polynomial regression, use `ggplot2::geom_smooth` directly (C)    


```{r example-regression}
plotObject <-
  plotPredVsObs(
    data = data,
    mapping = aes(
      x = Obs,
      y = Pred,
      groupby = Sex
    ),
    metaData = metaData,
    addRegression = TRUE
  ) +
  labs(
    title = "regression line",
    tag = "A"
  )

plot(plotObject)


plotObject +
  ggpubr::stat_regline_equation(aes(label = after_stat(rr.label))) +
  labs(
    title = "with RR as label",
    tag = "B"
  )


# C Local Polynomial Regression Fitting
plotPredVsObs(
  data = data,
  mapping = aes(
    x = Obs,
    y = Pred,
    groupby = Sex
  ),
  metaData = metaData,
  addRegression = FALSE
) +
  geom_smooth(
    method = "loess",
    formula = "y~x",
    na.rm = TRUE
  ) +
  labs(
    title = "Local Polynomial Regression Fitting",
    tag = "C"
  )
```


## 2.4 Add Guest-criteria lines

To compare DDI-ratios set the variable `addGuestLimits` to TRUE and set variable `deltaGuest`:

```{r predVsObs_Guest,fig.width=6,fig.height=6}
plotPredVsObs(
  data = DDIdata,
  mapping = aes(
    x = Obs,
    y = Pred,
    groupby = Study
  ),
  metaData = DDImetaData,
  addGuestLimits = TRUE,
  comparisonLineVector = list(identity = 1),
  deltaGuest = 1
)
```


## 2.5 Use non square format

Per default `plotPredVsObs()` produces a square plot with aspect ratio 1 and the same limits for x and y axis.
In the example below the Predicted values are set to 1/2 of the original values. A square plot does not make sense any more.
There for the variable `asSquarePlot` is set to `FALSE`.

```{r example-nonsquare}
DataNonSquare <- data %>%
  dplyr::mutate(Pred = Pred / 2)

plotPredVsObs(
  data = DataNonSquare,
  mapping = aes(
    x = Obs,
    y = Pred,
    groupby = Sex
  ),
  metaData = metaData,
  asSquarePlot = FALSE
)
```

# 3. Residuals vs covariate  (`plotResVsCov()`)

## 3.1 Basic examples

The function `plotResVsCov()` calculates the residuals using the aesthetics `observed` and `predicted`.


### 3.1.1 Default settings

The default value for the variable `residualScale` is "log". Residuals will then be calculated by $log(observed) - log(predicted)$.
A horizontal comparison line with value 0 is displayed.
Aesthetic `groupby` can be used to group observations.

```{r reVsObs-default}
plotResVsCov(
  data = data,
  mapping = aes(
    x = Age,
    predicted = Pred,
    observed = Obs,
    groupby = Sex
  )
)
```

## 3.1.2 Linear scale for residuals

If the variable `residualScale` is set to "linear",  residuals will  be calculated by $observed - predicted$.

Below the `linetype` of the comparison line was set to 'solid'


```{r resVsCov-basic-linear}
plotResVsCov(
  data = data, mapping = aes(
    x = Age,
    predicted = Pred,
    observed = Obs,
    groupby = Sex
  ),
  residualScale = "linear",
  geomComparisonLineAttributes = list(linetype = "solid")
)
```

## 3.1.3 Map residuals directly 

It is also possible to use pre-calculated residuals or formulas by using the aestehtic 'y'

```{r resVsCov-basic-mappedToY}
plotResVsCov(data = data, mapping = aes(
  x = Age,
  y = Obs - Pred,
  groupby = Sex
))
```

## 3.2 Adjusting Comparsion Lines


```{r resVsCov-comparisonlines}
plotResVsCov(
  data = data,
  mapping = aes(
    x = Age,
    predicted = Pred,
    observed = Obs,
    groupby = Sex
  ),
  comparisonLineVector = list(zero = 0, "lower limit" = -0.25, "upper limit" = 0.25)
)
```




## 3.3 Adding a regression line

To add a regression line set the input variable `addRegression` to `TRUE`.
For regression lines the package {ggpubr} has a nice function `stat_regline_equation` to add statistics of the regression as labels 

```{r, warning=FALSE, fig.width=8}
plotResVsCov(
  data = data,
  mapping = aes(
    x = Age,
    predicted = Pred,
    observed = Obs,
    groupby = Sex
  ),
  addRegression = TRUE
) +
  ggpubr::stat_regline_equation(aes(label = after_stat(eq.label)))
```

# 4. Ratio plots (`plotRatioVsCov()`)

## 4.1 Basic examples

### 4.1.1 Default settings

`plotRatioVsCov()` is used to evaluate ratios vs a covariate. 
As default as comparison the identity and 1.5 and 2 point lines are added, default for `yscale` is 'log'.
Aesthetic `groupby` can be used to group observations.

```{r ratio-defaults}
plotRatioVsCov(
  data = pkRatioData,
  mapping = aes(
    x = Age,
    y = Ratio,
    groupby = Sex
  ),
  metaData = metaData
)
```

### 4.1.2 Compare residuals as ratio

Within the function `plotRatioVsCov()` the variable `residualScale` is fixed to "ratio", the ratio of the residuals is then calculated as $observed/predicted$.
Below the comparison line was set to a 1.2 fold distance.


```{r ratio-residuals}
plotRatioVsCov(
  data = data, mapping = aes(
    x = Age,
    predicted = Pred,
    observed = Obs,
    groupby = Sex
  ),
  comparisonLineVector = getFoldDistanceList(c(1.2))
)
```
### 4.1.3 Using {ospsuite.plot} specific `aesthetics` like mdv and `error_relative`

If some of the data should be omitted, we could do this by mapping a `logcial` column to the aesthetic `mdv`. 
Below we excluded data with Age less than 20.

Additional error bars are displayed by mapping the column gsd  to the aesthetic `error_relative`.



```{r ratio-error}
plotRatioVsCov(
  data = pkRatioData,
  mapping = aes(
    x = Age,
    y = Ratio,
    error_relative = gsd,
    mdv = Age < 20,
    groupby = Sex
  )
) +
  theme(
    legend.box = "horizontal",
    legend.title = element_blank()
  )
```


## 4.2. Qualification of Ratios

If the package data.table is installed and the variable `comparisonLineVector` is a named list, the {ggplot} object returned by `plotRatioVsCov` has an additional entry `countsWithin`,  which contains a `data.frame` with the Ratios within the specific ranges given by variable `comparisonLineVector`.

To calculate this the function `getCountsWithin` is called internally without stratification (variable `groups= NULL`).
If you want to calculate the Number of Ratios  with stratification you can call `getCountsWithin` and define the stratification with the variable `groups`


```{r qualification-plot, results='asis'}
plotObject <- plotRatioVsCov(
  data = pkRatioData,
  mapping = aes(
    x = Age,
    y = Ratio,
    groupby = Sex
  ),
  metaData = metaData
)


plot(plotObject)
```


To compare DDI-ratios set the variable `addGuestLimits` to TRUE and set variable `deltaGuest`:



```{r quaification-tables, results='asis'}
knitr::kable(plotObject$countsWithin, caption = "Ratios not stratified")


CountsWithin_Sex <- getCountsWithin(
  data = pkRatioData,
  yColumn = "Ratio",
  groups = c("Sex")
)

knitr::kable(CountsWithin_Sex, caption = "Ratios stratified by gender")
```


## 4.3 Add Guest-criteria lines


```{r ratio-Guest}
plotRatioVsCov(
  data = DDIdata,
  mapping = aes(
    x = Obs,
    predicted = Pred,
    observed = Obs,
    groupby = Study
  ),
  metaData = DDImetaData,
  addGuestLimits = TRUE,
  comparisonLineVector = 1,
  deltaGuest = 1
)
```



# 5. qq-plot (`plotQQ()`)



`plotQQ()` produces a Quantile-quantile plot.
 If using the aesthetics `predicted` and `observed`, residuals will be calculated according to the variable `residualScale`.
 In the example below, default value log is used.


```{r qq-log}
plotQQ(
  data = data,
  mapping = aes(
    predicted = Pred,
    observed = Obs,
    groupby = Sex
  )
)
```
The aesthetic 'sample' columns with pre-calculated residuals or formulas can be mapped directly

```{r qqplot}
plotQQ(
  data = data,
  mapping = aes(
    sample = Obs - Pred,
    groupby = Sex
  ),
  residualScale = "linear"
)
```

# 6 Usage of aesthetics `predicted` and `observed` in other functions

The aesthetics `observed` and `predicted` can also be used in the functions `plotHistogram()` and  `plotBoxWhisker()`
The residuals will be calculated as defined by variable `residualScale`.


## 6.1 Residuals as histogram


```{r histogram-residuals}
plotHistogram(
  data = data,
  metaData = metaData,
  mapping = aes(
    predicted = Pred,
    observed = Obs,
    groupby = Sex
  ),
  residualScale = "log",
  plotAsFrequency = TRUE,
  distribution = "normal"
) +
  geom_vline(xintercept = 0, linetype = "dashed")
```

## 6.2 Stratify residuals with a boxwhisker plot

```{r boxwhisker-residuals}
plotBoxWhisker(
  mapping = aes(
    predicted = Pred,
    observed = Obs,
    x = Sex
  ),
  data = pkRatioData,
  metaData = metaData,
  residualScale = "linear"
)
```




```{r cleanup,echo=FALSE}
resetDefaults(oldDefaults)
```
