---
title: "Histogram Plots"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Histogram Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 300,
  fig.show = "hold",
  fig.align = "center",
  fig.width = 7,
  out.width = "80%",
  comment = "#>"
)
```


# 1. Introduction

This vignette documents and illustrates workflows for producing histograms using the function `plotHistogram` from the `ospsuite.plots` package.

## 1.1 Setup

This vignette uses the `{ospsuite.plots}` and `{tidyr}` libraries. We will use the default settings of `{ospsuite.plots}` (see vignette("ospsuite.plots", package = "ospsuite.plots")) but will adjust the legend position.

```{r setup}
library(ospsuite.plots)
library(tidyr)

# Set Defaults
oldDefaults <- ospsuite.plots::setDefaults()

# Place default legend position above the plot for clearer histogram plots
theme_update(legend.position = "top")
theme_update(legend.direction = "horizontal")
theme_update(legend.title = element_blank())
```

## 1.2 Example Data

This vignette uses the following datasets:

- **Data Set 1**:

```{r load-data-1, results='asis'}
histData <- exampleDataCovariates %>%
  dplyr::filter(SetID == "DataSet1") %>%
  dplyr::select(c("ID", "Sex", "Age", "AgeBin", "Ratio"))

# Metadata
metaData <- attr(exampleDataCovariates, "metaData")
metaData <- metaData[intersect(names(histData), names(metaData))]

knitr::kable(head(histData), digits = 2, caption = "First rows of example data.")
knitr::kable(metaData2DataFrame(metaData), digits = 2, caption = "List of meta data")
```

- **Data Set 2**:

```{r load-data-2, results='asis'}
histDataDistr <- exampleDataCovariates %>%
  dplyr::filter(SetID == "DataSet2") %>%
  dplyr::select(c("ID", "AgeBin", "Sex", "Obs"))

# Metadata for Distribution Data
metaDataDistr <- attr(exampleDataCovariates, "metaData")
metaDataDistr <- metaDataDistr[intersect(names(histDataDistr), names(metaDataDistr))]

knitr::kable(head(histDataDistr), digits = 2, caption = "First rows of distribution data.")
knitr::kable(metaData2DataFrame(metaDataDistr), digits = 2, caption = "List of meta data for distribution data")
```

# 2. Examples

## 2.1 Illustration of Basic Histograms

### 2.1.1 Basic Example

Histogram of the "Ratio" column mapped to `x`, stratified by the "Sex" column mapped to `fill`.

```{r minimal-example-default}
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, fill = Sex),
  metaData = metaData
)
```

### 2.1.2 Basic Example: Change of Defaults

The variable `geomHistAttributes` is set by default to `getDefaultGeomAttributes("Hist")`, which is a list with entries `bins = 10` and `position = ggplot2::position_nudge()`.

In the example below, the variable `geomHistAttributes` is set to a list with entry `position = "dodge"`. This changes the position, but note that the default value of `geomHistAttributes` contains the entry `bins = 10`, which is now overwritten, and the default `{ggplot}` number of 30 is used.

```{r minimal-example-position, warning=FALSE, message=FALSE}
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, groupby = Sex),
  metaData = metaData,
  geomHistAttributes = list(position = "dodge")
)
```

### 2.1.3 Basic Example: Change of Position but Keep Number of Bins

To preserve the default settings, we modified the variable with `utils::modifyList(getDefaultGeomAttributes("Hist"), list(position = "stack"))`. This changes the position but preserves the number of bins.

```{r minimal-example-stack}
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, groupby = Sex),
  metaData = metaData,
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Hist"),
    list(position = "stack")
  )
)
```

### 2.1.4 Basic Example: Overlay of Histograms

By setting the position to `identity` and setting `alpha` to a value below 1, an overlay of histograms is produced.

```{r minimal-example-overlay}
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, fill = Sex),
  metaData = metaData,
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Hist"),
    list(position = "identity", binwidth = 1, alpha = 0.5)
  )
)
```

### 2.1.5 Omit Data Points Flagged as Missing Dependent Variable (MDV)

If some of the data should be omitted, we can do this by mapping a boolean to the aesthetic `mdv`. Below, we exclude data above the value of 4:

```{r minimal-example-mdv}
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, fill = Sex, mdv = Ratio > 4),
  metaData = metaData
)
```

### 2.1.6 Stratified by a Combination of Columns

To stratify by a combination of columns, use the function `interaction` for the mapping to `groupby`:

```{r minimal-example-interaction}
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, groupby = interaction(Sex, AgeBin, sep = "-")),
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Hist"),
    list(position = "dodge")
  ),
  metaData = metaData
)
```

### 2.1.7 Customization of Binning

Use the input variable `geomHistAttributes` to change the binning. The entries of this list are passed to `ggplot2::geom_histogram`, which provides many possibilities to customize the binning. Below, we define the bin boundaries by adding the entry `breaks` to `geomHistAttributes`.

```{r minimal-example-binboundaries}
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, fill = Sex),
  geomHistAttributes = list(position = position_nudge(), breaks = seq(0, 5, 0.5)),
  metaData = metaData
)
```

You could also map a binning function to the aesthetic `x`. Below, `ggplot2::cut_number` is used to create 3 bins with equal numbers of observations. The data is now displayed as categorical data.

```{r minimal-example-binfunction}
plotHistogram(
  data = histData,
  mapping = aes(x = cut_number(Ratio, n = 3, labels = c("low", "mean", "high")), fill = Sex),
  geomHistAttributes = list(position = position_nudge()),
  metaData = metaData
) + labs(x = "Ratio")
```

## 2.2 Frequency

If the variable `plotAsFrequency` is set to TRUE and:

- `position` is `stack`: frequency is calculated as count per bin / total (A)
- `position` is NOT `stack`: frequency is calculated as count per bin / per group (B)

```{r examples-frequency}
# A
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, groupby = Sex),
  metaData = metaData,
  plotAsFrequency = TRUE,
  geomHistAttributes = list(bins = 10, position = "stack")
) + labs(tag = "A", caption = "Frequency is calculated as count per bin / total")

# B
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, groupby = Sex),
  metaData = metaData,
  plotAsFrequency = TRUE
) + labs(tag = "B", caption = "Frequency is calculated as count per bin / per group")
```

Both plots could also be calculated by directly setting `y` in the mapping:

- `position` is `stack`: frequency is calculated as count per bin / total (C)
- `position` is NOT `stack`: frequency is calculated as count per bin / per group (D)

```{r examples-frequency-manual}
# C
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, fill = Sex, y = after_stat(count / sum(count))),
  metaData = metaData,
  plotAsFrequency = FALSE,
  geomHistAttributes = list(bins = 10, position = "stack")
) + labs(tag = "C", caption = "Frequency is calculated as count per bin / total")

# D
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, fill = Sex, y = after_stat(density)),
  metaData = metaData,
  plotAsFrequency = FALSE
) + labs(tag = "D", caption = "Frequency is calculated as count per bin / per group")
```

# 3. Distribution Fit

The optional input variable `distribution` provides the possibility of fitting the data distribution. All distributions from the package `{stats}` are available (see `?stats::distributions`). Internally, `ggh4x::stat_theodensity` is used for the fit. Check the help for more details.

For the most common distributions, the keys "normal" (instead of `norm`) and "lognormal" (instead of `lnorm`) are also accepted.

The vertical line indicates the mean. The function to calculate the mean is determined by the input variable `meanFunction`. Available options are:

- `none` (no line is plotted)
- `mean` (arithmetic mean)
- `geomean` (geometric mean)
- `median`
- `auto` (default, selects the mean function according to the selected distribution)

Below are examples for:

### 3.1 Fit of a Normal Distribution with Mean as Vertical Line

```{r example-distribution}
# Plot normal distribution
plotHistogram(
  data = histDataDistr,
  mapping = aes(x = Obs, fill = Sex),
  metaData = metaDataDistr,
  distribution = "normal"
)
```

### 3.2 Fit of a Chi-Squared Distribution without Vertical Line

```{r example-distribution-chisq}
plotHistogram(
  data = histDataDistr,
  mapping = aes(x = Obs, groupby = Sex),
  metaData = metaDataDistr,
  distribution = "chisq",
  meanFunction = "none"
)
```

### 3.3 Fit of Stacked Data

With the option `stack`, it is also possible to get the distribution of the sum only.

```{r example-stacked-distribution}
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, fill = Sex),
  metaData = metaData,
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Hist"),
    list(position = "stack")
  ),
  distribution = "normal"
)
```

### 3.4 Fit with Frequency TRUE

To fit a frequency, select a distribution (here "normal") and set the variable `plotAsFrequency` to TRUE.

```{r example-frequency-fit}
plotHistogram(
  data = histDataDistr,
  mapping = aes(x = Obs, fill = Sex),
  metaData = metaDataDistr,
  distribution = "normal",
  plotAsFrequency = TRUE
)
```

### 3.5 Fit with Frequency TRUE and Stacked Data

```{r example-frequency-stacked-fit}
plotHistogram(
  data = histData,
  mapping = aes(x = Ratio, fill = Sex),
  metaData = metaData,
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Hist"),
    list(position = "stack")
  ),
  distribution = "normal",
  plotAsFrequency = TRUE
)
```

### 3.6 X-Axis on Log Scale for Distribution Fit

As the fit is based on binning, and binning is dependent on scale, a log scale has to be set before the distribution fit. Please use the variable `xscale = 'log'` and do not add a `{ggplot}` like `scale_x_log10`.

```{r example-logscale}
plotHistogram(
  data = histDataDistr,
  mapping = aes(x = Obs, fill = Sex),
  metaData = metaDataDistr,
  xscale = "log",
  distribution = "norm",
  meanFunction = "none"
) + labs(tag = "A")
```

# 4. Histogram for Categorical Data

The function `plotHistogram` can also be used to plot categorical data with a bar plot. Internally, the function switches from `geom_histogram` to `geom_bar`. With default inputs, the function switches automatically to a bar plot if the data is a factor or non-numeric. (See plot A). It can also be done manually by setting the variable `asBarPlot` to TRUE (see plot B).

```{r example-categorical-data}
# A Input is factor
plotHistogram(
  data = histData,
  mapping = aes(x = AgeBin, fill = Sex),
  metaData = metaData
) + labs(tag = "A")

# B Set asBarPlot = TRUE to convert input to factor
plotHistogram(
  data = histData,
  mapping = aes(x = round(histData$Age / 10) * 10, fill = Sex),
  asBarPlot = TRUE,
  metaData = metaData
) + labs(x = "Age rounded to 10 years", tag = "B")
```

```{r cleanup, echo=FALSE}
resetDefaults(oldDefaults)
```

```
