---
title: "Histogram Plots"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Histogram Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 300,
  fig.show = "hold",
  fig.align = "center",
  fig.width = 7,
  out.width = "80%",
  comment = "#>"
)
```


# 1. Introduction

The following vignette aims at documenting and illustrating workflows for producing histograms using the function `plotHistogram` from the `ospsuite.plots` package.


## 1.1 Setup

For this vignette we will use the libraries {ospsuite.plots} and {tidyr}.
We use the default settings of {ospsuite.plots} (see vignette("ospsuite.plots", package = "ospsuite.plots")) but adjust the legend position

```{r setup}
library(ospsuite.plots)
library(ggplot2)
library(tidyr)

# setDefaults
oldDefaults <- ospsuite.plots::setDefaults()


# Place default legend position above the plot for prettier time profile plots
theme_update(legend.position = "top")
theme_update(legend.direction = "horizontal")
theme_update(legend.title = element_blank())
```


## 1.2 Example data

In this vignette following Data sets are used:

* Data set 1

```{r load-data-1, results='asis'}
histData <- exampleDataCovariates %>%
  dplyr::filter(SetID == "DataSet1") %>%
  dplyr::select(c("ID", "Sex", "Age", "AgeBin", "Ratio"))

metaData <- attr(exampleDataCovariates, "metaData")
metaData <- metaData[intersect(names(histData), names(metaData))]


# histData
knitr::kable(utils::head(histData), digits = 2, caption = "data")

knitr::kable(metaData2DataFrame(metaData), digits = 2, caption = "list of meta data")
```

* Data Set 2


```{r load-data-2, results='asis'}
histDataDistr <- exampleDataCovariates %>%
  dplyr::filter(SetID == "DataSet2") %>%
  dplyr::select(c("ID", "AgeBin", "Sex", "Obs"))

metaDataDistr <- attr(exampleDataCovariates, "metaData")
metaDataDistr <- metaData[intersect(names(histData), names(metaDataDistr))]


# histData
knitr::kable(utils::head(histDataDistr), digits = 2, caption = "first rows of data")

knitr::kable(metaData2DataFrame(metaDataDistr), digits = 2, caption = "list of meta data")
```

# Examples

## 2.1 Illustration of basic histograms


### 2.1.1 Basic example 

Histogram of column "Ratio" mapped to `x` stratified by column "Sex" mapped to `fill`.

```{r minimalExample_default}
plotHistogram(
  data = histData,
  mapping <- aes(
    x = Ratio,
    fill = Sex
  ),
  metaData = metaData
)
```

### 2.1.2 Basic example: change of defaults

The variable `geomHistAttributes`  is set per default to `getDefaultGeomAttributes("Hist")`
which is a list with entries `bins = 10` and `position = ggplot2::position_nudge()`

In the example below the variable `geomHistAttributes` is set to a list with entry `position = "dodge"`
That changes the position, but attention, the default value of `geomHistAttributes` contains the entry `bins = 10` which is now overwritten and the default {ggplot} number of 30 is used.

The column "Sex" is here mapped to `groupby`. `groupby` is an additional aesthetics of {ospsuite.plots} which works together with the variable `groupAesthetics`. For the function `plotHistogram()`  `groupAesthetics` is not settable and fixed to `fill`.


```{r minimalExample_position, warning=FALSE,message=FALSE}
plotHistogram(
  data = histData,
  mapping <- aes(
    x = Ratio,
    groupby = Sex
  ),
  metaData = metaData,
  geomHistAttributes = list(position = "dodge")
)
```


### 2.1.3 Basic example: change of position but keep number of bins

To preserve the default settings, we modified the variable with `utils::modifyList(getDefaultGeomAttributes("Hist"), 
list(position = "stack"))` 

That changes the position but preserves the number of bins


```{r minimalExample_stack}
plotHistogram(
  data = histData,
  mapping <- aes(
    x = Ratio,
    groupby = Sex
  ),
  metaData = metaData,
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Hist"),
    list(position = "stack")
  )
)
```

### 2.1.4 Basic example: overlay of histograms

By setting the position to `identity` and setting `alpha` to a value below 1, an overlay of histograms is produced.

```{r minimalExample_overlay}
plotHistogram(
  data = histData,
  mapping <- aes(
    x = Ratio,
    groupby = Sex
  ),
  metaData = metaData,
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Hist"),
    list(
      position = "identity",
      binwidth = 1,
      alpha = 0.5
    )
  )
)
```

### 2.1.5 Omit data points flagged as Missing Dependent Variable (mdv)

If some of the data should be omitted, we could do this by mapping a boolean to the aesthetic `mdv`. Below we excluded data above the value of 4: 


```{r minimalExample_mdv}
plotHistogram(
  data = histData,
  mapping <- aes(
    x = Ratio,
    fill = Sex,
    mdv = Ratio > 4
  ),
  metaData = metaData
)
```


### 2.1.6 Stratified by a combination of columns

To stratify by a combination of columns use function `interaction` for the mapping to `groupby`:

```{r minimalExample_interaction}
plotHistogram(
  data = histData,
  mapping <- aes(
    x = Ratio,
    groupby = interaction(Sex, AgeBin, sep = "-")
  ),
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Hist"),
    list(position = "dodge")
  ),
  metaData = metaData
)
```

### 2.1.7 Customization of binning

Use the input variable `geomHistAttributes` to change the binning. The entries of this list are passed to
`ggplot2::geom_histogram`, which provides many possibilities to customize the binning. Below we defined the bin boundaries 
by adding the entry `breaks` to `geomHistAttributes`.  
(See `?geom_histogram` for more details)

```{r minimalExample_binboundaries}
plotHistogram(
  data = histData,
  mapping <- aes(
    x = Ratio,
    groupby = Sex,
  ),
  geomHistAttributes =
    list(
      position = position_nudge(),
      breaks = seq(0, 5, 0.5)
    ),
  metaData = metaData
)
```

You could also map a binning function to the aesthetic x. Below `ggplot2::cut_number` is used to create 3 bins with equal numbers of observations.
Th data is now displayed as categorical data (see section 4)


```{r minimalExample_binfunction}
plotHistogram(
  data = histData,
  mapping <- aes(
    x = cut_number(Ratio,
      n = 3,
      labels = c("low", "mean", "high")
    ),
    groupby = Sex,
  ),
  geomHistAttributes =
    list(
      position = position_nudge()
    ),
  metaData = metaData
) +
  labs(x = "Ratio")
```

## 2.2. Frequency

If variable `plotAsFrequency` is set to TRUE and 

- `position` is `stack`   frequency is calculated as count per bin / total (A)
- `position` is NOT `stack`   frequency is calculated as count per bin / per group (B)



```{r examples-frequency}
# A)
plotHistogram(
  data = histData,
  mapping = aes(
    x = Ratio,
    groupby = Sex,
  ),
  metaData = metaData,
  plotAsFrequency = TRUE,
  geomHistAttributes = list(bins = 10, position = "stack")
) +
  labs(
    tag = "A",
    caption = "frequency is calculated as count per bin / total"
  )

# B)
plotHistogram(
  data = histData,
  mapping = aes(
    x = Ratio,
    groupby = Sex,
  ),
  metaData = metaData,
  plotAsFrequency = TRUE
) +
  labs(
    tag = "B",
    caption = "frequency is calculated as count per bin / per group"
  )
```


both plots could also be calculated by directly setting y in the mapping

- `position` is `stack`   frequency is calculated as count per bin / total (C)
- `position` is NOT `stack`   frequency is calculated as count per bin / per group (D)


```{r examples-frequency_manual}
# C)
plotHistogram(
  data = histData,
  mapping = aes(
    x = Ratio,
    fill = Sex,
    y = after_stat(count / sum(count))
  ),
  metaData = metaData,
  plotAsFrequency = FALSE,
  geomHistAttributes = list(bins = 10, position = "stack")
) +
  labs(
    tag = "c",
    caption = "frequency is calculated as count per bin / total"
  )


# D)
plotHistogram(
  data = histData,
  mapping = aes(
    x = Ratio,
    fill = Sex,
    y = after_stat(density)
  ),
  metaData = metaData,
  plotAsFrequency = FALSE
) +
  labs(
    tag = "D",
    caption = "frequency is calculated as count per bin / per group"
  )
```


# 3. Distribution fit

The optional input variable `distribution` aims at providing the possibility of fitting the data distribution.  

All distributions from package {stats}  are available (see `? stats::distributions`).  
Internally `ggh4x::stat_theodensity` is used for the fit. Check the help for more details.

For the most common distributions the key "normal" (instead of `norm`) and "lognormal" (instead of `lnorm`) are also accepted.  


The vertical line indicates the mean. The function to calculate the mean is determined by the input variable `meanFunction`.   
Available are  

* `none`  (no line is plotted)
* `mean`  (arithmetic mean)
* `geomean`  (geometric mean) 
* `median`  
* `auto`  This is the default. It selects the mean function according to the selected distribution:   

   * `mean` for `normal` distribution  
   * `geomean` for `lognormal` distribution
   * `median` for all other distributions
   * `none`  if the distribution is also set to `none`.  
   

Below examples are shown for 

### 3.1 Fit of a `normal` distribution with `mean` as vertical line

```{r example-distribution}
# Plot normal distribution
plotHistogram(
  data = histDataDistr,
  mapping = aes(
    x = Obs,
    fill = Sex,
  ),
  metaData = metaDataDistr,
  distribution = "normal"
)
```


### 3.2 Fit of a `chisq` distribution without vertical line


```{r example-distribution-chisq}
plotHistogram(
  data = histDataDistr,
  mapping = aes(
    x = Obs,
    groupby = Sex,
  ),
  metaData = metaDataDistr,
  distribution = "chisq",
  meanFunction = "none"
)
```


## 3.3 Fit of stacked data

With option `stack`, it is also possible to get the distribution of the sum only.

```{r example-2-distributions-stack}
plotHistogram(
  data = histData,
  mapping = aes(
    x = Ratio,
    fill = Sex,
  ),
  metaData = metaData,
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Hist"),
    list(position = "stack")
  ),
  distribution = "normal"
)
```



## 3.2 Fit with frequency TRUE

To fit a frequency select a distribution (here "normal") and set variable `plotAsFrequency` to `TRUE` 

```{r example-frequency-fit}
plotHistogram(
  data = histDataDistr,
  mapping = aes(
    x = Obs,
    fill = Sex,
  ),
  metaData = metaDataDistr,
  distribution = "normal",
  plotAsFrequency = TRUE
)
```


## 3.3 Fit with frequency TRUE and stacked data


```{r example-frequency-stacked}
plotHistogram(
  data = histData,
  mapping = aes(
    x = Ratio,
    fill = Sex,
  ),
  metaData = metaData,
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Hist"),
    list(position = "stack")
  ),
  distribution = "normal",
  plotAsFrequency = TRUE,
)
```





## 3.4 x-axis on log scale for distribution fir

As fit is based on binning, and binning is depended on scale, a log scale has to be set before the distribution fit. Please use variable `xscale = 'log'` and do not add a {ggplot} like `scale_x_log10`

```{r example-logscale}
plotHistogram(
  data = histDataDistr,
  mapping = aes(
    x = Obs,
    fill = Sex,
  ),
  metaData = metaDataDistr,
  xscale = "log",
  distribution = "norm",
  meanFunction = "none"
) +
  labs(tag = "A")
```



# 4 histogram for categorical data

The function `plotHistogram` can also be used to plot categorical data with a bar plot.
Internally the function switches from geom_histogram to geom_bar.
With default inputs the functions switches automatically to bar plot, if the data is a factor or non numeric. (See plot A)
It can be done also manual by setting the variable `asBarPlot` to TRUE (see plot B)


```{r example-categorical-data}
# A  Input is factor
plotHistogram(
  data = histData,
  mapping <- aes(
    x = AgeBin,
    fill = Sex
  ),
  metaData = metaData
) +
  labs(tag = "A")


# B set asBarPlot = TRUE to convert input to factor
plotHistogram(
  data = histData,
  mapping <- aes(
    x = round(histData$Age / 10) * 10,
    fill = Sex
  ),
  asBarPlot = TRUE,
  metaData = metaData
) +
  labs(
    x = "age rounded to 10 years",
    tag = "B"
  )
```

```{r cleanup,echo=FALSE}
resetDefaults(oldDefaults)
```

