---
title: "Histogram Plots"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Histogram Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ospsuite.plots)
library(tidyr)

# setDefaults
ospsuite.plots::setDefaults()


# Place default legend position above the plot for prettier time profile plots
theme_update(legend.position = "top")
theme_update(legend.title = element_blank())
```

# 1. Introduction

The following vignette aims at documenting and illustrating workflows for producing histograms using the function `plotHistogram` from the `ospsuite.plots` package.

## 1.1 Example data

In this vignette following Data sets are used:

###  Data set 1

```{r load-data-1, results='asis'}
histData <- exampleDataCovariates %>%
  dplyr::filter(SetID == "DataSet1") %>%
  dplyr::select(c("ID", "Sex", "Age", "AgeBin", "Ratio"))

metaData <- attr(exampleDataCovariates, "metaData")
metaData <- metaData[intersect(names(histData), names(metaData))]


# histData
knitr::kable(utils::head(histData), digits = 2, caption = "data")

knitr::kable(metaData2DataFrame(metaData), digits = 2, caption = "list of meta data")
```

### Data Set 2


```{r load-data-2, results='asis'}
histData_distr <- exampleDataCovariates %>%
  dplyr::filter(SetID == "DataSet2") %>%
  dplyr::select(c("ID", "AgeBin", "Sex", "Obs"))

metaData_distr <- attr(exampleDataCovariates, "metaData")
metaData_distr <- metaData[intersect(names(histData), names(metaData_distr))]


# histData
knitr::kable(utils::head(histData_distr), digits = 2, caption = "first rows of data")

knitr::kable(metaData2DataFrame(metaData_distr), digits = 2, caption = "list of meta data")
```


# 2. Illustration of basic histograms


## 2.1 Examples for different bins and position

Attention the defaults for the variable `geomHistAttributes` is a list with entries bins = 10 and  position = ggplot2::position_nudge().
So if you change this variable and you do not want to change binning or position use `utils::modifyList` as in example C below
`geomHistAttributes` accepts all arguments `ggplot2::geom_histogram` accepts.

```{r minimalExample_bins, warning=FALSE,message=FALSE}
mapping <- aes(
  x = Ratio,
  fill = Sex
)

# A Defaults
plotHistogram(
  data = histData,
  mapping = mapping,
  metaData = metaData
) +
  labs(
    title = "position_nudge(), 10 bins (default)",
    tag = "A"
  )


# B ) position dodge default bins
plotHistogram(
  data = histData,
  mapping = mapping,
  metaData = metaData,
  geomHistAttributes = list(position = "dodge")
) +
  labs(
    title = "position dodge, 30 bins",
    tag = "B"
  )

# C) add position to default input for geomHistAttributes (nBins = 10)
plotHistogram(
  data = histData,
  mapping = mapping,
  metaData = metaData,
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Bar"),
    list(position = "stack")
  )
) +
  labs(
    title = "position stack , 10 bins",
    tag = "C"
  )

# add position and alpha to default input for geomHistAttributes (binwidth 1)
plotHistogram(
  data = histData,
  mapping = mapping,
  metaData = metaData,
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Bar"),
    list(
      position = "identity",
      binwidth = 1,
      alpha = 0.5
    )
  )
) +
  labs(
    title = "position identity, with opacity , binwidth",
    tag = "D"
  )
```

## 2.2. Frequency

If variable `plotAsFrequency` is set to TRUE and 

- `position` is `stack`   frequency is calculated as count per bin / total (A)
- `position` is NOT `stack`   frequency is calculated as count per bin / per group (B)

both plot could also be calculated by directly setting y in the mapping

- `position` is `stack`   frequency is calculated as count per bin / total (C)
- `position` is NOT `stack`   frequency is calculated as count per bin / per group (D)


```{r}
# A)
plotHistogram(
  data = histData,
  mapping = mapping,
  metaData = metaData,
  plotAsFrequency = TRUE,
  geomHistAttributes = list(bins = 10, position = "stack")
) +
  labs(
    tag = "A",
    caption = "frequency is calculated as count per bin / total"
  )

# B)
plotHistogram(
  data = histData,
  mapping = mapping,
  metaData = metaData,
  plotAsFrequency = TRUE
) +
  labs(
    tag = "B",
    caption = "frequency is calculated as count per bin / per fill"
  )


# C)
plotHistogram(
  data = histData,
  mapping = aes(
    x = Ratio,
    fill = Sex,
    y = after_stat(count / sum(count))
  ),
  metaData = metaData,
  plotAsFrequency = FALSE,
  geomHistAttributes = list(bins = 10, position = "stack")
) +
  labs(
    tag = "c",
    caption = "frequency is calculated as count per bin / total"
  )



# D)
plotHistogram(
  data = histData,
  mapping = aes(
    x = Ratio,
    fill = Sex,
    y = after_stat(density)
  ),
  metaData = metaData,
  plotAsFrequency = FALSE
) +
  labs(
    tag = "D",
    caption = "frequency is calculated as count per bin / per group"
  )
```


# 3. Distribution fit


## 3.1 Fit of different distributions and mean functions

The optional input `distribution` aims at providing the possibility of fitting the data distribution.
All distributions from package {stats}  are available (see ? stats::distributions).
For the most common the key normal (instead of norm) and lognormal (instead of `lnorm`) are also accepted.
As vertical line a mean is plotted. The function to calculate the mean is determined by the variable `meanFunction`. Available are `none`,`mean`,`geomean`,`median` and `auto`.
`auto` is the default is selects the mean function according to the selected distribution: `mean` for `normal` distribution, `geomean` for `lognormal` distribution, `median` for all other distributions, and `none`  if the distribution is also set to `none`.
Internally `ggh4x::stat_theodensity` is used for the fit. Check the help for more details

Below examples are shown for 

- A)  `normal` distribution with mean
- B)  `chisq` distribution without mean

```{r example-distribution}
# Plot normal distribution
plotHistogram(
  data = histData_distr,
  mapping = aes(
    x = Obs,
    fill = Sex,
  ),
  metaData = metaData_distr,
  distribution = "normal"
) +
  labs(tag = "A")



# Plot normal distribution
plotHistogram(
  data = histData_distr,
  mapping = aes(
    x = Obs,
    fill = Sex,
  ),
  metaData = metaData_distr,
  distribution = "chisq",
  meanFunction = "none"
) +
  labs(tag = "B")
```


## 3.2 Fit of stacked data



With option `stack`, it is also possible to get the distribution of the sum only while splitting the content of the bars.

```{r example-2-distributions-stack}
# Plot normal distribution of sum but bars are split by gender
plotHistogram(
  data = histData,
  mapping = aes(
    x = Ratio,
    fill = Sex,
  ),
  metaData = metaData,
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Bar"),
    list(position = "stack")
  ),
  distribution = "normal"
) +
  labs(tag = "A")
```



## 3.2 Fit with frequency TRUE

```{r}
# Plot normal distribution
plotHistogram(
  data = histData_distr,
  mapping = aes(
    x = Obs,
    fill = Sex,
  ),
  metaData = metaData_distr,
  distribution = "normal",
  plotAsFrequency = TRUE
) +
  labs(tag = "A")


# Plot normal distribution of sum but bars are split by gender
plotHistogram(
  data = histData,
  mapping = aes(
    x = Ratio,
    fill = Sex,
  ),
  metaData = metaData,
  geomHistAttributes = utils::modifyList(
    getDefaultGeomAttributes("Bar"),
    list(position = "stack")
  ),
  distribution = "normal",
  plotAsFrequency = TRUE,
) +
  labs(tag = "B")
```





## 3.3 x-axis on log scale for distribution fir

As fit is based on binning, and binning is depended on scale, a log scale has to be set before the distribution fit. Please use variable `xscale`

```{r}
plotHistogram(
  data = histData_distr,
  mapping = aes(
    x = Obs,
    fill = Sex,
  ),
  metaData = metaData_distr,
  xscale = "log",
  distribution = "norm",
  meanFunction = "none"
) +
  labs(tag = "A")
```



# 4 histogram for categorical data

The function `plotHistogram` can also be used to plot categorical data with a bar plot.
Internally the function switches from geom_histogram to geom_bar.
With default inputs the functions switches automatically to bar plot, if the data is a factor or non numeric. (See plot A)
It can be done also manual by setting the variable `asBarPlot` to TRUE (see plot B)




```{r}
# A Defaults
plotHistogram(
  data = histData,
  mapping = mapping <- aes(
    x = AgeBin,
    fill = Sex
  ),
  metaData = metaData
) +
  labs(tag = "A")


# A Defaults
plotHistogram(
  data = histData,
  mapping = mapping <- aes(
    x = round(histData$Age / 10) * 10,
    fill = Sex
  ),
  asBarPlot = TRUE,
  metaData = metaData
) +
  labs(
    x = "age rounded to 10 years",
    tag = "B"
  )
```

