---
title: "Overview"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{ospsuite.plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 300,
  fig.show = "hold",
  fig.align = "center",
  fig.width = 7,
  out.width = "80%",
  comment = "#>"
)

options(rmarkdown.html_vignette.check_title = FALSE)

library(ospsuite.plots)
library(tidyr)
```

# 1. Introduction

The main purpose of the `ospsuite.plots` library is to provide standardized plots typically used in the context of PBPK modeling. The library supports plot generation for the packages `OSPSuiteR` and `OSPSuite.ReportingEngine`.

The library is based on `ggplot2` functionality and also utilizes the `ggh4x` package.

# 2. Styling ospsuite.plots Plots

Every `ospsuite.plots` plot function already produces a fully styled plot, so you do not need to configure anything to get the `ospsuite.plots` look. When you *do* want to change something, there are two distinct mechanisms, and it helps to know which one to reach for:

- **OSP theme and custom scales** ([Section 2.1](#osp-theme-and-custom-scales)) shape the *frame and the mapped aesthetics* of a single plot, by adding them like any other `ggplot2` component: `theme_osp()` for the theme (background, grid, legend, titles), and `scale_colour_osp()` / `scale_fill_osp()` / `scale_shape_osp()` for the discrete `ospsuite.plots` color and shape palettes. They return ordinary `ggplot2` objects and never change global state, so they also let you bring the `ospsuite.plots` look to a plot you built yourself with plain `ggplot2`.
- **Session options** ([Section 2.2](#session-options)) set *defaults read at plot-function call time*: the geom aesthetics of the drawn marks (fill alpha, line width, error-bar caps, ...), statistical behaviour (percentiles, histogram bins), and side features (watermark, export format). Set them once and every later `ospsuite.plots` plot in the session picks them up; or pass the matching `geom*Attributes` argument to override just one call.

In short: reach for a **constructor** to restyle the look of *one plot* (especially a hand-built `ggplot2` plot); set an **option** to change the *default marks or behaviour* across your analysis.

Examples within this vignette are plotted using the following test data:

```{r get-test-data}
testData <- exampleDataCovariates |>
  dplyr::filter(SetID == "DataSet1") |>
  dplyr::select(c("ID", "Age", "Obs", "Pred", "Sex"))
```

## 2.1 OSP Theme and Custom Scales

The OSP theme and scales are added to a plot like any other `ggplot2` component and return ordinary `ggplot2` objects, so they do not change global state. The `ospsuite.plots` plot functions apply them internally; you add them yourself only when styling a plain `ggplot2` plot.

The same plot with the default `ggplot2` layout (A) and with the `ospsuite.plots` constructors applied (B):

```{r default-layout-comparison, warning=FALSE, fig.show='hold', fig.alt="Two scatter plots of the same data. Plot A uses the default ggplot2 layout with the grey theme and default colors. Plot B applies the ospsuite.plots constructors theme_osp() and scale_colour_osp(), showing the white theme and ospsuite.plots colors."}
basePlot <- ggplot(testData, aes(x = Obs, y = Pred, color = Sex)) +
  geom_point()

# A: default ggplot2 layout
basePlot + labs(title = "A: default ggplot2")

# B: ospsuite.plots constructors added to the same plot
basePlot +
  theme_osp() +
  scale_colour_osp() +
  labs(title = "B: ospsuite.plots (theme_osp + scale_colour_osp)")
```

### 2.1.1 Theme: `theme_osp()`

`theme_osp()` is a regular `ggplot2` theme, built on `theme_bw()`, that styles the **frame** of a plot (panel background, grid, legend placement, title and subtitle); it controls everything that is not the plotted data. Because it is an ordinary theme object, it behaves exactly like the built-in `theme_*()` themes: use it as the starting point and layer your own choices on top with a `theme()` call (the later `theme()` wins for the elements it sets). It also accepts the usual `theme_bw()` arguments such as `base_size`.

```{r theme_osp, warning=FALSE, fig.alt="Scatter plot of the ggplot2 mpg dataset (engine size versus highway mileage, coloured by car class) styled by starting from theme_osp() and overriding it: the legend is moved to the left, the title and subtitle are left-aligned and the major grid is dotted."}
# start from the OSP theme, then apply custom choices on top (these win)
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  labs(
    title = "Highway mileage vs engine size",
    subtitle = "ggplot2 mpg dataset, coloured by car class",
    x = "engine displacement (L)",
    y = "highway miles per gallon"
  ) +
  theme_osp() +
  theme(
    legend.position = "left",
    plot.title = element_text(hjust = 0),
    plot.subtitle = element_text(hjust = 0),
    panel.grid.major = element_line(linetype = "dotted")
  )
```

### 2.1.2 Color: `scale_colour_osp()` and `scale_fill_osp()`

`scale_colour_osp()` and `scale_fill_osp()` apply the `ospsuite.plots` discrete color palette to the mapped `colour`/`fill` aesthetics (`colorMaps$default` for up to 6 groups, `colorMaps$ospDefault` beyond; see `?colorMaps`). As in `ggplot2`, `scale_color_osp()` is an alias of `scale_colour_osp()`. As ordinary scales, they are overridden by any other `scale_colour_*()` / `scale_fill_*()` you add.

The palette used beyond six groups (`colorMaps$ospDefault`):

```{r scale-fill-osp, echo=FALSE, fig.asp = 1, fig.width = 7, out.width = "85%", fig.alt="Color swatch grid of the ospsuite.plots ospDefault palette, the colors used by scale_fill_osp() and scale_colour_osp() beyond six groups."}
scales::show_col(colorMaps$ospDefault, labels = FALSE, borders = "white")
```

### 2.1.3 Shape: `scale_shape_osp()`

All point layers produced by `ospsuite.plots` use `geom_point_osp()`, which draws from the OSP shape palette defined in `ospShapeNames`. When you build your own plots and want them to combine consistently with `ospsuite.plots` outputs, prefer `geom_point_osp()` over the raw `ggplot2::geom_point()` (raw `geom_point()` keeps using ggplot2's native pch shapes and is unaffected). Shape ordering is controlled per plot:

- `scale_shape_osp()` assigns shapes automatically from `ospShapeNames`.
- `scale_shape_osp_manual(values = c(level1 = "circle", level2 = "diamond", ...))` sets an explicit mapping.
- `scale_shape_osp_identity()` is the right choice when the data already contains shape names.

The available shapes:

```{r default-shapes, echo=FALSE, out.width = "90%", fig.asp = 0.7, fig.width = 8, fig.alt="Chart displaying all default OSP shape types available in ospsuite.plots. Shows various point shapes arranged in a grid with labels, demonstrating the visual appearance of each shape option."}
shapes <- data.frame(
  shapeNames = ospShapeNames,
  x = rep(1:6, length.out = length(ospShapeNames)),
  y = rep(1:4, each = 6, length.out = length(ospShapeNames))
)

ggplot(shapes, aes(x, y, shape = shapeNames)) +
  geom_point_osp(size = 5, color = "blue", fill = "red") +
  scale_shape_osp_identity() +
  geom_text(aes(label = shapeNames), nudge_y = -0.3, size = 3) +
  theme_void()
```

### 2.1.4 Applying the constructors globally (deprecated)

To style many unrelated (non-`ospsuite.plots`) plots in a session at once, `setDefaultTheme()` sets `theme_osp()` as the global `ggplot2` theme via `theme_set()`, and `setDefaults()` additionally mutates the global geom defaults and discrete color options. Both are **deprecated**: `ospsuite.plots` plots no longer need them, and the per-plot `theme_osp()` and scales above are the recommended path. They still work and return the previous settings so you can restore them with `resetDefaultTheme()` / `resetDefaults()`.

```{r defaultTheme, warning=FALSE}
# Deprecated: style every plot in the session, including non-ospsuite.plots ones
oldTheme <- ospsuite.plots::setDefaultTheme()

# Restore the previous global theme
ospsuite.plots::resetDefaultTheme(oldTheme)
```

## 2.2 Session Options

Session options are read by the `ospsuite.plots` plot functions at call time (falling back to the package defaults when an option is unset), so setting one changes the default for every later plot in the session. Set them with `options(ospsuite.plots.<key> = value)` and read them with `getOption("ospsuite.plots.<key>")`; the examples below use this form. The helpers `setOspsuite.plots.option(optionKey, value)` and `getOspsuite.plots.option(optionKey)` do the same using the keys from the `OptionKeys` enumeration, and `getDefaultOptions()` returns the full default list.

```{r, eval=FALSE}
# inspect the default options
ospsuite.plots::getDefaultOptions()
```

The options fall into three groups:

1. **Geom default aesthetics** ([Section 2.2.1](#geom-default-aesthetics)): how the drawn marks look (fill alpha, line width, error-bar caps, bar outline, ...).
2. **Statistical and plotting behaviour** ([Section 2.2.2](#statistical-and-plotting-behaviour)): what is computed or drawn, not how it looks (percentiles, histogram bins, LLOQ handling).
3. **Side features** ([Section 2.2.3](#side-features-watermark-and-export)): the watermark and the export format.

### 2.2.1 Geom Default Aesthetics

All plot functions have `geom*Attributes` input variables that are passed on to the corresponding `ggplot2` layer; their defaults are session options. This is how the `ospsuite.plots` look of the drawn marks (such as the fill alpha of ribbons and boxes, the line width, or the error-bar cap width) is applied per layer, rather than by mutating the global `ggplot2` geom defaults. Override one for a single plot by passing the argument (e.g. `plotRangeDistribution(..., geomRibbonAttributes = list(alpha = 0.3))`), or for the session by setting the option.

```{r options_defaultgeoms, echo=FALSE}
df <- rbind(
  data.frame(
    defaultOption = "Line",
    inputVariable = c("geomLineAttributes"),
    functions = c("plotTimeProfile()")
  ),
  data.frame(
    defaultOption = "Ribbon",
    inputVariable = "geomRibbonAttributes",
    functions = c("plotTimeProfile()")
  ),
  data.frame(
    defaultOption = "Point",
    inputVariable = "geomPointAttributes",
    functions = c(
      "plotTimeProfile()",
      "plotResVsCov()",
      "plotRatioVsCov()",
      "plotPredVsObs()"
    )
  ),
  data.frame(
    defaultOption = "Errorbar",
    inputVariable = "geomErrorbarAttributes",
    functions = c(
      "plotTimeProfile()",
      "plotResVsCov()",
      "plotRatioVsCov()",
      "plotPredVsObs()"
    )
  ),
  data.frame(
    defaultOption = "LLOQ",
    inputVariable = "geomLLOQAttributes",
    functions = c(
      "plotTimeProfile()",
      "plotResVsCov()",
      "plotRatioVsCov()",
      "plotPredVsObs()"
    )
  ),
  data.frame(
    defaultOption = "Hist",
    inputVariable = "geomHistAttributes",
    functions = c("plotHistogram()")
  ),
  data.frame(
    defaultOption = "Boxplot",
    inputVariable = c("geomBoxplotAttributes", "geomPointAttributes"),
    functions = c("plotBoxWhisker()", "plotBoxWhisker()")
  ),
  data.frame(
    defaultOption = "ComparisonLine",
    inputVariable = c("geomComparisonLineAttributes"),
    functions = c("plotResVsCov()", "plotRatioVsCov()", "plotPredVsObs()")
  ),
  data.frame(
    defaultOption = "GuestLine",
    inputVariable = c("geomGuestLineAttributes"),
    functions = c("plotResVsCov()", "plotRatioVsCov()", "plotPredVsObs()")
  )
) |>
  data.table::setDT() |>
  (\(x) {
    x[,
      .(inputVariable = paste(inputVariable, collapse = ", ")),
      by = c("defaultOption", "functions")
    ]
  })() |>
  tidyr::pivot_wider(
    names_from = defaultOption,
    values_from = inputVariable,
    values_fill = " "
  )

knitr::kable(df, align = "c", caption = "Usage of options for geom attributes")
```

The default fill alpha and line width are shared across these attributes through the `alpha` option (default `0.5`); setting it adjusts the transparency of every filled mark:

```{r geom-alpha-option, eval=FALSE}
# Make all filled marks more transparent for the session
options(ospsuite.plots.alpha = 0.3)
```

### 2.2.2 Statistical and Plotting Behaviour

These options change *what* a plot computes or draws, not how the marks look.

- `percentiles`: five quantiles (lower whisker, lower box, median, upper box, upper whisker) used by `plotBoxWhisker()` (default `c(0.05, 0.25, 0.5, 0.75, 0.95)`).
- `defaultPercentiles`: three quantiles used by `plotRangeDistribution()` (default `c(0.05, 0.5, 0.95)`).
- `geomHistAttributes$bins`: the default number of histogram bins.
- `lloqAlphaVector` / `lloqLineType`: the transparency of points below the LLOQ and the line type of the LLOQ line.

```{r behaviour-options, eval=FALSE}
# Change box-whisker percentiles for the session
options(ospsuite.plots.percentiles = c(0.1, 0.25, 0.5, 0.75, 0.9))

# Change range-plot percentiles for the session
options(ospsuite.plots.defaultPercentiles = c(0.1, 0.5, 0.9))
```

### 2.2.3 Side Features: Watermark and Export

#### Watermark

All plots in this package are created with `ggplotWithWatermark()` instead of `ggplot()`, which adds a watermark on print. The watermark is **enabled by default**; setting the option to `NULL` is treated the same as the default, so it still appears. You only need these options to customize or remove it.

```r
# Disable watermarks globally (e.g. in your .Rprofile, editable via usethis::edit_r_profile())
options(ospsuite.plots.watermarkEnabled = FALSE)
```

**Attention!** If you combine plots, e.g. with `cowplot::plot_grid`, the default print function without watermark is called. In that case add the watermark with `addWatermark(plotObject)` before printing.

The watermark is controlled by three options:

- `watermarkEnabled` (default `TRUE`): switch the watermark on or off.
- `watermarkLabel` (default `"preliminary analysis"`): the text.
- `watermarkFormat` (default `list(x = 0.5, y = 0.5, color = "lightgrey", angle = 30, fontsize = 12, alpha = 0.7)`): the appearance.

```{r default-watermark, fig.alt="Plot with the default ospsuite.plots watermark: light grey text reading 'preliminary analysis' at a 30-degree angle in the centre of the panel."}
# A: default watermark
ggplotWithWatermark() + labs(title = "A: default watermark")
```

```{r customize-watermark, fig.alt="Plot demonstrating a customized watermark: a large, bold, semi-transparent purple 'CONFIDENTIAL' label rotated 45 degrees across the centre of the panel."}
# B: customized watermark
options(ospsuite.plots.watermarkLabel = "CONFIDENTIAL")
options(
  ospsuite.plots.watermarkFormat = list(
    x = 0.5,
    y = 0.5,
    color = "purple",
    angle = 45,
    fontsize = 18,
    alpha = 0.35
  )
)

ggplotWithWatermark() + labs(title = "B: customized watermark")
```

```{r disable-watermark, fig.alt="Plot with the watermark disabled: a clean blank coordinate system with no watermark text overlay."}
# C: no watermark
options(ospsuite.plots.watermarkEnabled = FALSE)

ggplotWithWatermark() + labs(title = "C: no watermark")
```

```{r reset-watermark-default, include=FALSE}
# restore defaults for the rest of the vignette
options(ospsuite.plots.watermarkFormat = NULL)
options(ospsuite.plots.watermarkLabel = NULL)
options(ospsuite.plots.watermarkEnabled = TRUE)
```

#### Export

`exportPlot()` (see [Section 5](#plot-export)) uses these options as defaults, passed on to `ggplot2::ggsave()`:

- `exportWidth`: width of the exported file
- `exportUnits`: units for width and height
- `exportDevice`: device for plot export
- `exportDpi`: plot resolution

# 3. Plot Functions

All plot functions listed below call internally the function `initializePlot()`. This function constructs labels from the metadata and adds a watermark layer. It can also be used to create a customized ggplot.

## 3.1 `plotTimeProfile()`

[Time Profile Plots](plot-time-profile.html)

## 3.2 `plotBoxWhisker()`

[Box Whisker Plots](box-whisker-vignette.html)

## 3.3 `plotHistogram()`

[Histogram Plots](histogram.html)

## 3.4 `plotPredVsObs()`

[Goodness of Fit](Goodness_of_fit.html)

## 3.5 `plotResVsCov()`

[Goodness of Fit](Goodness_of_fit.html)

## 3.6 `plotRatioVsCov()`

[Goodness of Fit](Goodness_of_fit.html)

## 3.7 `plotQQ()`

[Goodness of Fit](Goodness_of_fit.html)

# 4. Additional Aesthetics

This package provides some additional aesthetics.

For more details, see the examples within the vignettes for the respective functions.

- `groupby`: Shortcut to use different aesthetics to group. All functions where this aesthetic is used have also a variable `groupAesthetics`. The mapping `groupby` is copied to all aesthetics listed within this variable and to the aesthetic `group`.
- `lloq`: Mapped to a column with values indicating the lower limit of quantification, adds horizontal (or vertical) lines to the plot. All observed values below the "lloq" are plotted with a lighter alpha. As values are compared row by row, it is possible to have more than one LLOQ.
- `error`: Mapped to a column with additive error (e.g., standard deviation); error bars are plotted. This is a shortcut to map `ymin` and `ymax` directly (`ymin = y - error` and `ymax = y + error`). Additionally, if `yScale` is set, `ymin` values below 0 are set to `y`.
- `error_relative`: Mapped to a column with relative error (e.g., geometric standard deviation); error bars are plotted. This is a shortcut to map `ymin` and `ymax` directly (`ymin = y / error_relative` and `ymax = y * error_relative`).
- `y2axis`: Creates a plot with 2 y axes. It is used to map to a column with a logical value. Values where this column has a TRUE entry will be displayed with a secondary axis.
- `mdv`: Mapped to a logical column. Rows where this column has entries set to TRUE are not plotted (MDV = missing data value, taken from NONMEM notation).
- `observed` / `predicted`: For the function `plotPredVsObs()`, `observed` is mapped to `x` and `predicted` is mapped to `y`.

Residuals and ratios must be pre-calculated before passing them to the plotting functions. If you are using the `{ospsuite}` package, use `ospsuite::addResidualColumn()` to add a residual column to your data.

See `vignette("Goodness of Fit", package = "ospsuite.plots")` for examples.

```{r additional-aesthetic-table, echo=FALSE}
df <- rbind(
  data.frame(
    aesthetic = "`groupby`",
    functions = c(
      "plotTimeProfile()",
      "plotHistogram()",
      "plotPredVsObs()",
      "plotResVsCov()",
      "plotRatioVsCov()",
      "plotQQ()"
    )
  ),
  data.frame(
    aesthetic = "`lloq`",
    functions = c("plotTimeProfile()", "plotPredVsObs()")
  ),
  data.frame(
    aesthetic = "`error`",
    functions = c(
      "plotTimeProfile()",
      "plotRatioVsCov()",
      "plotPredVsObs()",
      "plotResVsCov()"
    )
  ),
  data.frame(
    aesthetic = "`error_relative`",
    functions = c(
      "plotTimeProfile()",
      "plotRatioVsCov()",
      "plotPredVsObs()",
      "plotResVsCov()"
    )
  ),
  data.frame(aesthetic = "`y2axis`", functions = "plotTimeProfile()"),
  data.frame(
    aesthetic = "`mdv`",
    functions = c(
      "plotTimeProfile()",
      "plotBoxWhisker()",
      "plotHistogram()",
      "plotPredVsObs()",
      "plotResVsCov()",
      "plotRatioVsCov()",
      "plotQQ()"
    )
  ),
  data.frame(
    aesthetic = "`observed` / `predicted`",
    functions = c("plotPredVsObs()")
  )
) |>
  dplyr::mutate(used = "X") |>
  tidyr::pivot_wider(
    names_from = aesthetic,
    values_from = used,
    values_fill = " "
  )

knitr::kable(
  df,
  align = "c",
  caption = "Applicability of additional aesthetics in functions"
)
```

# 5. Plot Export

This section demonstrates how to use the `exportPlot` function to save `ggplot` objects to files, adjusting the width and height of the exported plots as necessary. This function is part of the `ospsuite.plots` package and simplifies the process of exporting plots for various purposes, such as publication or presentation.

## 5.1 Basic Usage

To export a plot, you need a ggplot object. Here's a basic example:

Create a simple ggplot object:

```{r export-object, eval=FALSE}
plotObject <- ospsuite.plots::plotHistogram(
  data = testData,
  mapping = aes(x = Age)
)
```

Exporting the Plot: Using the `exportPlot` function, you can easily save this plot to a file:

Replace `"path/to/save"` with the actual directory path where you want to save the plot, and adjust the width and height parameters as needed.

```{r export-basic, eval=FALSE}
exportPlot(
  plotObject = plotObject,
  filepath = "path/to/save",
  filename = "myplot",
  width = 10,
  height = 8
)
```

## 5.2 Advanced Usage

### Adjusting Plot Dimensions Based on Content

The `exportPlot` function can automatically adjust the plot dimensions based on its content, such as the presence of a legend or the aspect ratio.

Assuming `plotObject` is your ggplot object:

```{r export-advanced, eval=FALSE}
exportPlot(
  plotObject = plotObject,
  filepath = "path/to/save",
  filename = "adjusted_plot.png"
)
```

In this case, you don't need to specify the width and height explicitly; the function calculates them for you. The default width value saved in the `ospsuite.plots` option `exportWidth` is used. If an aspect ratio is defined in the theme of the plot, height will be adjusted accordingly; otherwise, the function exports square figures.
