---
title: "ospsuite.plots"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{ospsuite.plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.show = "hold"
)
```

```{r setup}
library(ospsuite.plots)
library(tidyr)
```


# 1. Introduction

## 1.1. Objectives of ospsuite.plots


The main purpose of the `ospsuite.plots`-library is to provide standardized plots typically used in PBPK context.
`ospsuite.plots` supports the plot generation of the packages `OSPSuiteR`  and `OSPSuite.ReportingEngine`. 

The library is based on `ggplot2` functionality. It uses also the libraries `ggnewscale` and `ggh4xs`.

# 2. Default settings for Layout

`ospsuite.plots` provides default settings for the layout including theme, geometric aesthetics, color and shapes for distinct scales.


```{r get-test-data, echo=FALSE, results='asis'}
testData <- exampleDataCovariates %>%
  dplyr::filter(SetID == "DataSet1") %>%
  dplyr::select(c("ID", "Age", "Obs", "Pred", "Sex"))


knitr::kable(head(testData), digits = 3)
```




## 2.1 Plots with and without default layout

## 2.1.1 Default ggplot layout

- A) plot by ospsuite.plots function
- B) customized plot

```{r,warning=FALSE}
# ospsuite.plots function
plotHistogram(
  data = testData,
  mapping = aes(x = Age)
) +
  labs(tag = "A")

# customized plot

ggplot(
  data = testData,
  mapping = aes(x = Obs, y = Pred, color = Sex, shape = Sex)
) +
  geom_point() +
  theme(legend.position = "top") +
  labs(tag = "B")
```

## 2.1.2 set ospsuite.plots layout


```{r,warning=FALSE,fig.show='hold'}
# set default layout  and save previous layout in variable oldDefaults
oldDefaults <- ospsuite.plots::setDefaults()

# ospsuite.plots function
plotHistogram(
  data = testData,
  mapping = aes(x = Age)
) +
  labs(tag = "A")

# customized plot

ggplot(
  data = testData,
  mapping = aes(x = Obs, y = Pred, color = Sex, shape = Sex)
) +
  geom_point() +
  theme(legend.position = "top") +
  labs(tag = "B")
```

## 2.1.3 reset to previously save layout


```{r,warning=FALSE,fig.show='hold'}
# reset to previously saved layout options
ospsuite.plots::resetDefaults(oldDefaults = oldDefaults)

# ospsuite.plots function
plotHistogram(
  data = testData,
  mapping = aes(x = Age)
) +
  labs(tag = "A")

# customized plot

ggplot(
  data = testData,
  mapping = aes(x = Obs, y = Pred, color = Sex, shape = Sex)
) +
  geom_point() +
  theme(legend.position = "top") +
  labs(tag = "B")
```



## 2.2 Default Theme

Functions to set the ospsuite.plots default theme only are: `setDefaultTheme()` and `resetDefaultTheme()`. This functions are called by `setDefaults()` and `resetDefaults()`.

```{r}
# set ospsuite.plots Default theme
oldTheme <- ospsuite.plots::setDefaultTheme()

# customize theme using ggplot functionalities
theme_update(legend.position = "top")
theme_update(legend.title = element_blank())

# reset to the previously saved
ggplot2::theme_set(oldTheme)
```

## 2.3 Default Color

Functions to set the ospsuite.plots default color only are: `setDefaultColorMapDistinct()` and `resetDefaultColorMapDistinct()`. This functions are called by `setDefaults()` and `resetDefaults()`.

Colors are set to discrete and ordinal scales for `fill` and `colour`


```{r out.width="80%"}
# set ospsuite.plots Default Color
oldColors <- ospsuite.plots::setDefaultColorMapDistinct()


ggplot() +
  geom_tile(aes(
    x = rep(seq(1, 3), 2),
    y = rep(seq(1, 2), each = 3),
    fill = as.factor(seq(1, 6))
  )) +
  labs("up to 6 colors") +
  theme(
    legend.position = "none",
    axis.title = element_blank()
  )

ggplot() +
  geom_tile(aes(
    x = c(rep(seq(1, 7), 7), 1, 2),
    y = c(rep(seq(1, 7), each = 7), 8, 8),
    fill = as.factor(seq(1, 51))
  )) +
  labs("> 6 colors") +
  theme(
    legend.position = "none",
    axis.title = element_blank()
  )



# customize colors:
# set to gray colors
ospsuite.plots::setDefaultColorMapDistinct(ColorMaps[["grays"]])

ggplot() +
  geom_tile(aes(
    x = rep(seq(1, 3), 3),
    y = rep(seq(1, 3), each = 3),
    fill = as.factor(seq(1, 9))
  )) +
  theme(
    legend.position = "none",
    axis.title = element_blank()
  )

# set to Color palettes inspired by plots in Lancet journals
ospsuite.plots::setDefaultColorMapDistinct(ColorMaps[["ggsci::pal_lancet()"]])

ggplot() +
  geom_tile(aes(
    x = rep(seq(1, 3), 3),
    y = rep(seq(1, 3), each = 3),
    fill = as.factor(seq(1, 9))
  )) +
  theme(
    legend.position = "none",
    axis.title = element_blank()
  )



# reset to the previously saved
ospsuite.plots::resetDefaultColorMapDistinct(oldColorMaps = oldColors)
```


## 2.5 Default Options

`getDefaultOptions()` returns a list of options which are used in this package.
This Options are set by the function `setDefaults()` via the variable  `defaultOptions`.

```{r}
ospsuite.plots::setDefaults(defaultOptions = ospsuite.plots::getDefaultOptions())
```


## 2.5.1 Options to customize Watermark

Options to switch watermark on and off, select the label and customize format:
see also section [4. Watermark]

```{r,eval = FALSE}
options(list(
  ospsuite.plots.watermark_enabled = TRUE,
  ospsuite.plots.watermark_label = "preliminary analysis",
  ospsuite.plots.watermark_format = list(
    x = 0.5,
    y = 0.5,
    color = "grey20",
    angle = 30,
    fontsize = 12,
    alpha = 0.7
  )))
```

The watermark is added with function 
```{r}
plotObject <- ggplot() +
  layerWatermark()
```

This is done automatically by each plot function of this package



## 2.5.2 Options to set the defaults for geom layer attributes

```{r, eval = FALSE}
options(list(
  ospsuite.plots.geomLineAttributes = list(),
  ospsuite.plots.geomRibbonAttributes = list(color = NA),
  ospsuite.plots.geomPointAttributes = list(),
  ospsuite.plots.geomErrorbarAttributes = list(width = 0),
  ospsuite.plots.geomLLOQAttributes = list(linetype = "dotted"),
  ospsuite.plots.geomRatioLineAttributes = list(linetype = "dashed"),
  ospsuite.plots.geomBoxplotAttributes = list(
    position = position_dodge(width = 1),
    color = "black"
  ),
  ospsuite.plots.geomBarAttributes = list(
    bins = 10,
    position = ggplot2::position_nudge()
  )
))
```

## 2.5.3 Alpha

```{r, eval = FALSE}
# default alpha
options(ospsuite.plots.Alpha = 0.7)

# alpha of LLOQ values
options(ospsuite.plots.LLOQAlphaVector = c("TRUE" = 0.618, "FALSE" = 1))
```


## 2.5.3 Percentiles

Percentiles are used within the function `plotBoxwhisker`

```{r, eval = FALSE}
# percentiles for boxwhisker plots
options(ospsuite.plots.Percentiles = c(0.05, 0.25, 0.5, 0.75, 0.95))
```



# 3. Plot-functions

If a plot function is called  internally the function  `initializePlot` with variables `metaData` and `mapping` is called. This function constructs labels from the metadata, and adds a watermark layer.


## 3.1 `plotTimeProfile()`

see `\code{vignette("Time Profile Plots", package = "ospsuite.plots")}`

## 3.2 `plotBoxwhisker()`

not yet implemented

## 3.3 `plotCumulativeTimeProfile()`

not yet implemented

## 3.4 `plotHistogram`

see `\code{vignette("Histogram Plots", package = "ospsuite.plots")}`

## 3.5 `plotPkratio()`

not yet implemented

## 3.6 `plotDDiratio()`

not yet implemented

## 3.7 `plotObsVsPred()`

not yet implemented

## 3.8 `plotResVs()`

not yet implemented

## 3.9 `plotQQ()`

not yet implemented

## 3.10 `plotPiechart()`

not yet implemented



# 4. Shapes

## 4.1 Default shapes ggplot:


```{r, out.width = "90%", fig.asp = 0.4, fig.width = 8}
shape_names <- c(
  "circle", paste("circle", c("open", "filled", "cross", "plus", "small")), "bullet",
  "square", paste("square", c("open", "filled", "cross", "plus", "triangle")),
  "diamond", paste("diamond", c("open", "filled", "plus")),
  "triangle", paste("triangle", c("open", "filled", "square")),
  paste("triangle down", c("open", "filled")),
  "plus", "cross", "asterisk"
)

shapes <- data.frame(
  shape_names = shape_names,
  x = c(1:7, 1:6, 1:3, 5, 1:3, 6, 2:3, 1:3),
  y = -rep(1:6, c(7, 6, 4, 4, 2, 3))
)


ggplot(shapes, aes(x, y)) +
  geom_point(aes(shape = shape_names), color = "blue", fill = "red", size = 5, stroke = 1) +
  geom_text(aes(label = shape_names), nudge_y = -0.3, size = 3.5) +
  scale_shape_identity() +
  theme_void()
```

## 4.2 Use unicode symbols  with {showtext}

Attention 
- use of {showtext}  has side effects   
- a customized geom_Point function ha to be used

```{r,  fig.showtext=TRUE,out.width = "90%", fig.asp = 0.8, fig.width = 8}
shapes <- data.frame(
  shape_names = names(Shapes),
  shape_symbols = unlist(unname(Shapes)),
  x = rep(c(1:5), 8),
  y = rep(-c(1:8), each = 5)
)


# TODO functionName

showtext::showtext_auto()
ggplot(shapes, aes(x, y)) +
  geomPointUnicode(aes(shape = shape_symbols), color = "blue", fill = "red", size = 5) +
  geom_text(aes(label = shape_names), nudge_y = -0.3, size = 3.5) +
  scale_shape_identity() +
  theme_void()
showtext::showtext_auto(enable = "off")
```




