---
title: "ospsuite.plots"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{ospsuite.plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 300,
  fig.show = "hold",
  fig.align = "center",
  fig.width = 7,
  out.width = "80%",
  comment = "#>"
)
```

```{r setup}
library(ospsuite.plots)
library(tidyr)
```


# 1. Introduction

## 1.1. Objectives of ospsuite.plots


The main purpose of the `ospsuite.plots`-library is to provide standardized plots typically used in PBPK context.
`ospsuite.plots` supports the plot generation of the packages `OSPSuiteR`  and `OSPSuite.ReportingEngine`. 

The library is based on `ggplot2` functionality. It uses also the libraries `ggnewscale` and `ggh4xs`.

# 2. Default settings for Layout

`ospsuite.plots` provides default settings for the layout including theme, geometric aesthetics, color and shapes for distinct scales.

Examples within this vignette are plotted with following test data:

```{r get-test-data,  results='asis'}
testData <- exampleDataCovariates %>%
  dplyr::filter(SetID == "DataSet1") %>%
  dplyr::select(c("ID", "Age", "Obs", "Pred", "Sex"))


knitr::kable(head(testData), digits = 3)
```




## 2.1 Plots with and without default layout

## 2.1.1 Default ggplot layout

* A plot by ospsuite.plots function
* B customized plot

```{r,warning=FALSE}
# ospsuite.plots function
ospsuite.plots::plotHistogram(
  data = testData,
  mapping = aes(x = Age)
) +
  labs(tag = "A")

# customized plot

ggplot(
  data = testData,
  mapping = aes(x = Obs, y = Pred, color = Sex, shape = Sex)
) +
  geom_point() +
  theme(legend.position = "top") +
  labs(tag = "B")
```

## 2.1.2 set ospsuite.plots layout

To set the default we use the same logic as used in `ggplot2::theme_set()`. Invisibly the previous settings are returned so you can easily save it, then later restore it.

`setDefaults()` sets the theme, discrete color palette, shapes and various options. All objects can also be set separately as described below.


```{r,warning=FALSE,fig.show='hold'}
# set default layout  and save previous layout in variable old defaults
# inputs are set to default inputs
oldDefaults <- ospsuite.plots::setDefaults(
  defaultOptions = list(),
  ColorMapList = NULL,
  shapeValues = NULL,
  PointAsUnicode = FALSE
)

# ospsuite.plots function
ospsuite.plots::plotHistogram(
  data = testData,
  mapping = aes(x = Age)
) +
  labs(tag = "A")

# customized plot

ggplot(
  data = testData,
  mapping = aes(x = Obs, y = Pred, color = Sex, fill = Sex, shape = Sex)
) +
  geom_point() +
  theme(legend.position = "top") +
  labs(tag = "B")
```


## 2.1.3 reset to previously save layout


```{r,warning=FALSE,fig.show='hold'}
# reset to previously saved layout options
ospsuite.plots::resetDefaults(oldDefaults = oldDefaults)

# ospsuite.plots function
ospsuite.plots::plotHistogram(
  data = testData,
  mapping = aes(x = Age)
) +
  labs(tag = "A")

# customized plot

ggplot(
  data = testData,
  mapping = aes(x = Obs, y = Pred, color = Sex, shape = Sex)
) +
  geom_point() +
  theme(legend.position = "top") +
  labs(tag = "B")
```



## 2.2 Default Theme

Functions to set the ospsuite.plots default theme only are: `setDefaultTheme()` and `resetDefaultTheme()`. This functions are called by `setDefaults()` and `resetDefaults()`.

```{r}
# set ospsuite.plots Default theme
oldTheme <- ospsuite.plots::setDefaultTheme()

# customize theme using ggplot functionalities
theme_update(legend.position = "top")
theme_update(legend.title = element_blank())

# reset to the previously saved
resetDefaultTheme(oldTheme)
```

## 2.3 Default Color

Functions to set the ospsuite.plots default color only are: `setDefaultColorMapDistinct()` and `resetDefaultColorMapDistinct()`. This functions are called by `setDefaults()` and `resetDefaults()`.

Colors are set to discrete and ordinal scales for `fill` and `colour`

The package provides some color palettes in the object ColorMaps (See `? ColorMaps`)

The example below shows plots with  

* A plot with default settings for up to 6 different colors
* B plot with default settings for more the 6 different colors
* C customize settings for all following plots using the `ColorMaps[["grays"]]`
* D customize plot: set gray scale for this plot only using function `scale_fill_grey()`    

You can also use color Palettes from other packages e.g ggsci:  

* E customize settings for all following plots using as colormap ggsci::pal_lancet()(9)
* F customized plots set gray scale for this plot only using function `ggsci::scale_color_lancet()`    


```{r}
# set ospsuite.plots Default Color
oldColors <- ospsuite.plots::setDefaultColorMapDistinct()


ggplot() +
  geom_tile(aes(
    x = rep(seq(1, 3), 2),
    y = rep(seq(1, 2), each = 3),
    fill = as.factor(seq(1, 6))
  )) +
  labs(
    title = "default settings for up to 6 different colors",
    tag = "A"
  ) +
  theme(
    legend.position = "none",
    axis.title = element_blank()
  )
```


```{r}
ggplot() +
  geom_tile(aes(
    x = c(rep(seq(1, 7), 7), 1, 2),
    y = c(rep(seq(1, 7), each = 7), 8, 8),
    fill = as.factor(seq(1, 51))
  )) +
  labs(title = "default settings for more the 6 different colors") +
  theme(
    legend.position = "none",
    axis.title = element_blank()
  )
```


```{r}
# customize colors:
# set to gray colors
ospsuite.plots::setDefaultColorMapDistinct(ColorMaps[["grays"]])

ggplot() +
  geom_tile(aes(
    x = rep(seq(1, 3), 3),
    y = rep(seq(1, 3), each = 3),
    fill = as.factor(seq(1, 9))
  )) +
  theme(
    legend.position = "none",
    axis.title = element_blank()
  ) +
  labs(
    title = "ColorMaps gray",
    tag = "C"
  )
```


```{r}
ggplot() +
  geom_tile(aes(
    x = rep(seq(1, 3), 3),
    y = rep(seq(1, 3), each = 3),
    fill = as.factor(seq(1, 9))
  )) +
  theme(
    legend.position = "none",
    axis.title = element_blank()
  ) +
  scale_fill_grey() +
  labs(
    title = "scale_fill_grey",
    tag = "D"
  )
```


```{r}
# set to Color palettes inspired by plots in Lancet journals
ospsuite.plots::setDefaultColorMapDistinct(ggsci::pal_lancet()(9))

ggplot() +
  geom_tile(aes(
    x = rep(seq(1, 3), 3),
    y = rep(seq(1, 3), each = 3),
    fill = as.factor(seq(1, 9))
  )) +
  theme(
    legend.position = "none",
    axis.title = element_blank()
  ) +
  labs(
    title = "ggsci::pal_lancet",
    tag = "E"
  )
```


```{r}
# set to Color palettes inspired by plots in Lancet journals
ospsuite.plots::setDefaultColorMapDistinct(ggsci::pal_lancet()(9))

ggplot() +
  geom_tile(aes(
    x = rep(seq(1, 3), 3),
    y = rep(seq(1, 3), each = 3),
    fill = as.factor(seq(1, 9))
  )) +
  theme(
    legend.position = "none",
    axis.title = element_blank()
  ) +
  ggsci::scale_color_lancet() +
  labs(
    title = "ggsci::scale_color_lancet",
    tag = "F"
  )



# reset to the previously saved
ospsuite.plots::resetDefaultColorMapDistinct(oldColorMaps = oldColors)
```

## 2.4 Default Shapes

Functions to set the ospsuite.plots default shapes only are: `oldShapeValue = setDefaultShapeDistinct(shapeValues)` and `resetDefaultShapeDistinct(oldShapeValue)`. This functions are called by `setDefaults()` and `resetDefaults()`.
The scales are set to the option  `ospsuite.plots.shapeValues`, which is then used to
set the discrete scale of shapes for all {ospsuite.plots} functions.
for customized functions add `scale_shape_manual(values = getOption('ospsuite.plots.shapeValues'))`


## 2.5 Default Options

`getDefaultOptions()` returns a list of options which are used in this package.
This Options are set by the function `setDefaults()` via the variable  `defaultOptions`.

```{r}
ospsuite.plots::setDefaults(defaultOptions = ospsuite.plots::getDefaultOptions())
```


## 2.5.1 Options to customize Watermark

Per default a watermark is displayed in all plots

The watermark is added with function 
```{r}
plotObject <- ggplot() +
  layerWatermark()
```

This is done automatically by each plot function of this package


There are options to 

* switch watermark on and off  
* select the label 
* customize format


```{r,eval = FALSE}
options(list(
  ospsuite.plots.watermark_enabled = TRUE,
  ospsuite.plots.watermark_label = "preliminary analysis",
  ospsuite.plots.watermark_format = list(
    x = 0.5,
    y = 0.5,
    color = "grey20",
    angle = 30,
    fontsize = 12,
    alpha = 0.7
  )
))
```


examples to customize watermark  

* A change format of watermark
* B disabled watermark
* C reset to default

```{r}
# change format and label of watermark
options(list(
  ospsuite.plots.watermark_format = list(
    x = 0.2,
    y = 0.9,
    color = "red",
    angle = 90,
    fontsize = 24,
    alpha = 0.2
  ),
  ospsuite.plots.watermark_label = "NEW"
))

initializePlot() +
  labs(
    title = "changed watermark",
    tag = "A"
  )
```


disable watermark:

```{r}
options(list(ospsuite.plots.watermark_enabled = FALSE))

initializePlot() +
  labs(
    title = "no watermark",
    tag = "B"
  )
```

reset to default:

```{r}
options(list(
  ospsuite.plots.watermark_format = NULL,
  ospsuite.plots.watermark_label = NULL,
  ospsuite.plots.watermark_enabled = NULL
))

initializePlot() +
  labs(
    title = "default watermark",
    tag = "C"
  )
```



## 2.5.2 Options to set the defaults for geom layer attributes


```{r, eval = FALSE}
options(list(
  ospsuite.plots.geomLineAttributes = list(),
  ospsuite.plots.geomRibbonAttributes = list(color = NA),
  ospsuite.plots.geomPointAttributes = list(),
  ospsuite.plots.geomErrorbarAttributes = list(width = 0),
  ospsuite.plots.geomLLOQAttributes = list(linetype = "dotted"),
  ospsuite.plots.geomRatioLineAttributes = list(linetype = "dashed"),
  ospsuite.plots.geomBoxplotAttributes = list(
    position = position_dodge(width = 1),
    color = "black"
  ),
  ospsuite.plots.geomHistAttributes = list(
    bins = 10,
    position = ggplot2::position_nudge()
  ),
  ospsuite.plots.geomBarAttributes = list(
    position = ggplot2::position_nudge()
  )
))
```

## 2.5.3 Options to set defaults for aesthetic `alpha`

Options to set the face alpha of ribbons filled points   
and options to set the filled points for values below and above lloq

```{r, eval = FALSE}
# default alpha
options(ospsuite.plots.Alpha = 0.7)

# alpha of LLOQ values
options(ospsuite.plots.LLOQAlphaVector = c("TRUE" = 0.3, "FALSE" = 1))
```


## 2.5.3 Percentiles

Percentiles are used within the function `plotBoxwhisker()`

```{r, eval = FALSE}
# percentiles for boxwhisker plots
options(ospsuite.plots.Percentiles = c(0.05, 0.25, 0.5, 0.75, 0.95))
```



# 3. Plot-functions

All plot functions listed below call  internally the function  `initializePlot`. 
This function constructs labels from the metadata, and adds a watermark layer.
It can also be used to create a customized ggplot


## 3.1 `plotTimeProfile()`

see `vignette("Time Profile Plots", package = "ospsuite.plots")`

## 3.2 `plotBoxwhisker()`

see `vignette("Box-Whisker Plots", package = "ospsuite.plots")`

## 3.3 `plotHistogram`

see `vignette("Histogram Plots", package = "ospsuite.plots")`

## 3.4 `plotPredVsObs()`

see `vignette("Goodness of fit", package = "ospsuite.plots")`

## 3.5 `plotResVsCov()`

see `vignette("Goodness of fit", package = "ospsuite.plots")`

## 3.6 `plotRatioVsCov()`

see `vignette("Goodness of fit", package = "ospsuite.plots")`

## 3.7 `plotQQ()`

see `vignette("Goodness of fit", package = "ospsuite.plots")`


# 4. Additional aesthetics

This package provide some additional aesthetics.   

For more details see the example within the vignettes for the respective functions.


* `groupby`: Short cut to use different aesthetics to group.  
All function where this aesthetic is used have also a variable `groupAesthtics. The mapping `groupby`is copied to all aesthetics listed within this variable and to aesthetic `group`.  

* `lloq`: Mapped to column with value of lower limit of quantification, adds horizontal (respectively vertical) lines
to the plot. All observed values below the "lloq" are plotted with a lighter alpha.  
As values will be compared row by row, it is possible to have more then one lloq.   

* `error`: Mapped to column with additive error (e.g. standard deviation) errorbars are plotted. 
This is a short cut to map ymin and ymax directly (`ymin = y - error` and `ymax = y + error`).
Additionally if `yscale` is ymin values below 0 are set to y.

* `error_relative`: Mapped to column with relative error (e.g. geometric standard deviation) errorbars are plotted.
This is a short cut to map ymin and ymax directly (`ymin = y/error_relative` and `ymax = y*error_relative`). 

* `y2axis`: Creates a plot with 2 y axis. It is used to map to column with a logical.
Values where this column has entry TRUE will be displayed with a secondary axis.

* `mdv`: Mapped to a logical column. rows where this column entries are TRUE, are not plotted.
(mdv = missing data value is taken from NONMEM notation)

* `observed` / `predicted`: For the function `plotPredVsObs()` `observed` is mapped to `x`and `predicted` is mapped to y. 
For all other functions residuals are calculated if both aesthetics are mapped. 
The variable `residualScale` defines the calculation method: 
  * `residualScale` = "log" -> $log(observed) - log(predicted)$, 
  * `residualScale` = "linear" -> $observed - predicted$, 
  * `residualScale` = "ratio" -> $observed/predicted$  
see `vignette("Goodness of fit", package = "ospsuite.plots")` for examples


```{r additional-aesthetic-table, echo=FALSE}
df <- rbind(
  data.frame(
    aesthetic = "`groupby`",
    functions = c(
      "plotTimeProfile()",
      "plotHistogram()",
      "plotPredVsObs()",
      "plotResVsCov()",
      "plotRatioVsCov()",
      "plotQQ()"
    )
  ),
  data.frame(
    aesthetic = "`lloq`",
    functions = c(
      "plotTimeProfile()",
      "plotPredVsObs()"
    )
  ),
  data.frame(
    aesthetic = "`error`",
    functions = c(
      "plotTimeProfile()",
      "plotRatioVsCov()",
      "plotPredVsObs()",
      "plotResVsCov()"
    )
  ),
  data.frame(
    aesthetic = "`error_relative`",
    functions = c(
      "plotTimeProfile()",
      "plotRatioVsCov()",
      "plotPredVsObs()",
      "plotResVsCov()"
    )
  ),
  data.frame(
    aesthetic = "`y2axis`",
    functions = "plotTimeProfile()"
  ),
  data.frame(
    aesthetic = "`mdv`",
    functions = c(
      "plotTimeProfile()",
      "plotBoxwhisker()",
      "plotHistogram()",
      "plotPredVsObs()",
      "plotResVsCov()",
      "plotRatioVsCov()",
      "plotQQ()"
    )
  ),
  data.frame(
    aesthetic = "`observed` / `predicted`",
    functions = c(
      "plotBoxwhisker()",
      "plotHistogram()",
      "plotPredVsObs()",
      "plotResVsCov()",
      "plotRatioVsCov()",
      "plotQQ()"
    )
  )
) %>%
  dplyr::mutate(used = "X") %>%
  tidyr::pivot_wider(names_from = aesthetic, values_from = used, values_fill = " ")


knitr::kable(df, align = "c", caption = "Applicability of additional aesthetics in functions")
```


  



# 5. Shapes

## 5.1 Default shapes ggplot:

```{r, out.width = "90%", fig.asp = 0.4, fig.width = 8}
shape_names <- c(
  "circle", paste("circle", c("open", "filled", "cross", "plus", "small")), "bullet",
  "square", paste("square", c("open", "filled", "cross", "plus", "triangle")),
  "diamond", paste("diamond", c("open", "filled", "plus")),
  "triangle", paste("triangle", c("open", "filled", "square")),
  paste("triangle down", c("open", "filled")),
  "plus", "cross", "asterisk"
)


shapes <- data.frame(
  shape_names = shape_names,
  x = c(1:7, 1:6, 1:3, 5, 1:3, 6, 2:3, 1:3),
  y = -rep(1:6, c(7, 6, 4, 4, 2, 3))
)


ggplot(shapes, aes(x, y)) +
  geom_point(aes(shape = shape_names), color = "blue", fill = "red", size = 5, stroke = 1) +
  geom_text(aes(label = shape_names), nudge_y = -0.3, size = 3.5) +
  scale_shape_identity() +
  theme_void()
```

## 5.2 Use unicode symbols  with {showtext}

Attention 
- use of {showtext}  has side effects   
- a customized geom_Point function `geomPointUnicode` has to be used

```{r,  fig.showtext=TRUE,out.width = "90%", fig.asp = 0.8, fig.width = 8}
shapes <- data.frame(
  shape_names = names(Shapes),
  shape_symbols = unlist(unname(Shapes)),
  x = rep(c(1:5), 8),
  y = rep(-c(1:8), each = 5)
)


showtext::showtext_auto()
ggplot(shapes, aes(x, y)) +
  geomPointUnicode(aes(shape = shape_symbols), color = "blue", fill = "red", size = 5) +
  geom_text(aes(label = shape_names), nudge_y = -0.3, size = 3.5) +
  scale_shape_identity() +
  theme_void()
showtext::showtext_auto(enable = "off")
```


## 5.3 Switch between modes


To switch to the unicode mode call `setDefaults` with input variable `PointAsUnicode = TRUE`.
To switch back use `setDefaults(PointAsUnicode = FALSE)` or `resetDefaults(oldDefaults)`
But it is recommended to produce all plots of one workflow either with unicode mode or all plots without.
Other wise the plots may have different fonts.


```{r}
oldDefaults <- setDefaults(PointAsUnicode = TRUE)


dt <- data.frame(
  x = c(1, 2, 1, 2),
  y = c(1, 1, 2, 2),
  species = c("pig", "dog", "mouse", "rat")
)


plotObject <- plotYVsX(
  data = dt,
  mapping = aes(
    x = x,
    y = y,
    groupby = species
  ),
  xscale = "linear",
  xscale.args = list(limits = c(0.5, 2.5)),
  yscale = "linear",
  yscale.args = list(limits = c(0.5, 2.5))
)

plot(plotObject)
```


use non default icons

```{r}
plot(plotObject +
  scale_shape_manual(values = c(
    pig = "pig",
    dog = "dog",
    mouse = "mouse",
    rat = "rat"
  )))
```





```{r cleanup,echo=FALSE}
resetDefaults(oldDefaults)
```


