---
title: "Overview"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{ospsuite.plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 300,
  fig.show = "hold",
  fig.align = "center",
  fig.width = 7,
  out.width = "80%",
  comment = "#>"
)
```

```{r setup, echo=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)

library(ospsuite.plots)
library(tidyr)
```

# 1. Introduction

## 1.1 Objectives of ospsuite.plots

The main purpose of the `ospsuite.plots` library is to provide standardized plots typically used in the context of PBPK modeling. The library supports plot generation for the packages `OSPSuiteR` and `OSPSuite.ReportingEngine`.

The library is based on `ggplot2` functionality and also utilizes the `ggh4x` package.

# 2. Default Settings for Layout

`ospsuite.plots` provides default settings for the layout, including theme, geometric aesthetics, colors, and shapes for distinct scales.

Examples within this vignette are plotted using the following test data:

```{r get-test-data, results='asis'}
testData <- exampleDataCovariates |>
  dplyr::filter(SetID == "DataSet1") |>
  dplyr::select(c("ID", "Age", "Obs", "Pred", "Sex"))

knitr::kable(head(testData), digits = 3)
```

## 2.1 Plots with and without Default Layout

### 2.1.1 Default ggplot Layout

- A plot created using the `ospsuite.plots` function
- B customized plot

```{r default-layout-comparison, warning=FALSE, fig.alt="Two plots comparing default ggplot styling (A) versus ospsuite.plots styling (B). Plot A shows a histogram with default ggplot formatting. Plot B shows a scatter plot with customized colors, shapes, and legend positioning demonstrating enhanced styling capabilities."}
# ospsuite.plots function
ospsuite.plots::plotHistogram(data = testData, mapping = aes(x = Age)) + labs(tag = "A")

# Customized plot
ggplot(data = testData, mapping = aes(x = Obs, y = Pred, color = Sex, shape = Sex)) +
  geom_point() +
  theme(legend.position = "top") +
  labs(tag = "B")
```

### 2.1.2 Apply the ospsuite.plots Layout per Plot

The `ospsuite.plots` plot functions already apply the full `ospsuite.plots` layout (theme, colors, shapes, line width and other geom aesthetics) to each plot they produce, so their appearance does not depend on any global setting (plot A below).

For your own `ggplot2` plots, compose the per-plot constructors: `themeOspsuite()` for the theme and `scale_colour_osp()` / `scale_fill_osp()` for the discrete `ospsuite.plots` color palette. Shape ordering for OSP-aware geoms is controlled per plot via `scale_shape_osp()` or `scale_shape_osp_manual()`. None of these change the global `ggplot2` state (plot B below).

```{r ospsuite-plots-layout-per-plot, warning=FALSE, fig.show='hold', fig.alt="Two plots showing the ospsuite.plots layout applied per plot. Plot A shows a histogram from an ospsuite.plots function with the layout already applied. Plot B shows a scatter plot built with plain ggplot2 that adopts the ospsuite.plots theme and colors by adding themeOspsuite() and scale_colour_osp()/scale_fill_osp()."}
# ospsuite.plots function: layout already applied per plot
ospsuite.plots::plotHistogram(data = testData, mapping = aes(x = Age)) + labs(tag = "A")

# Your own ggplot2 plot: opt into the ospsuite.plots layout per plot
ggplot(data = testData, mapping = aes(x = Obs, y = Pred, color = Sex, fill = Sex, shape = Sex)) +
  geom_point() +
  themeOspsuite() +
  scale_colour_osp() +
  scale_fill_osp() +
  theme(legend.position = "top") +
  labs(tag = "B")
```

### 2.1.3 Setting the Layout Globally (deprecated)

`setDefaults()` mutated the global `ggplot2` state (theme, geom defaults and discrete color options) for the whole session. It is now **deprecated**: `ospsuite.plots` plots no longer need it, and the per-plot constructors above are the recommended way to style your own plots. `setDefaults()` (and `setDefaultTheme()` / `setDefaultColorMapDistinct()`) still work and remain available for styling many unrelated, non-`ospsuite.plots` plots in a session at once; the previous settings are returned invisibly so you can restore them with `resetDefaults()`.

## 2.2 Default Theme

The recommended way to apply the `ospsuite.plots` theme is per plot with `themeOspsuite()`, which returns a `ggplot2` theme object and does not change the global `ggplot2` state. The `ospsuite.plots` plot functions already apply it internally, so their styling is independent of the global theme.

```{r themeOspsuite}
# Apply the ospsuite.plots theme to a single plot
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  themeOspsuite()
```

For styling many unrelated plots at once, `setDefaultTheme()` and `resetDefaultTheme()` set or restore the `ospsuite.plots` theme as the global `ggplot2` theme for the whole session (`setDefaultTheme()` simply calls `theme_set(themeOspsuite())`). `setDefaultTheme()` is **deprecated** in favor of the per-plot `themeOspsuite()` shown above; it is kept only for the global-styling use case.

```{r defaultTheme, warning=FALSE}
# Set ospsuite.plots theme globally (deprecated; affects non-ospsuite.plots plots)
oldTheme <- ospsuite.plots::setDefaultTheme()

# Customize theme using ggplot functionalities
theme_update(legend.position = "top")
theme_update(legend.title = element_blank())

# Reset to the previously saved theme
ospsuite.plots::resetDefaultTheme(oldTheme)
```

## 2.3 Default Color

The recommended way to apply the `ospsuite.plots` discrete color palette is per plot, with `scale_colour_osp()` and `scale_fill_osp()`. They do not change the global `ggplot2` state and are applied automatically by the `ospsuite.plots` plot functions. The palette uses `colorMaps$default` for up to 6 groups and `colorMaps$ospDefault` beyond (see `?colorMaps`).

The example below shows plots with:

- A per-plot `ospsuite.plots` colors for up to 6 different colors (`scale_fill_osp()`)
- B per-plot `ospsuite.plots` colors for more than 6 different colors (`scale_fill_osp()`)
- C per-plot gray scale using ggplot2's `scale_fill_grey()`
- D per-plot Lancet palette using `ggsci::scale_fill_lancet()`

```{r default-color-settings-6colors, fig.alt="Tile plot demonstrating the ospsuite.plots color palette for up to 6 different colors. The plot shows a 3x2 grid of colored tiles, each a different discrete color from the default ospsuite.plots palette, applied per plot with scale_fill_osp()."}
ggplot() +
  geom_tile(aes(x = rep(seq(1, 3), 2), y = rep(seq(1, 2), each = 3), fill = as.factor(seq(1, 6)))) +
  scale_fill_osp() +
  labs(title = "ospsuite.plots colors for up to 6 different colors", tag = "A") +
  theme(legend.position = "none", axis.title = element_blank())
```

```{r default-color-settings-many-colors, fig.alt="Tile plot demonstrating the ospsuite.plots color palette for more than 6 different colors. The plot shows a 7x7 grid plus 2 additional tiles (51 total), illustrating how scale_fill_osp() extends to the larger ospDefault palette for many groups."}
ggplot() +
  geom_tile(aes(x = c(rep(seq(1, 7), 7), 1, 2), y = c(rep(seq(1, 7), each = 7), 8, 8), fill = as.factor(seq(1, 51)))) +
  scale_fill_osp() +
  labs(title = "ospsuite.plots colors for more than 6 different colors", tag = "B") +
  theme(legend.position = "none", axis.title = element_blank())
```

To use a different palette for a single plot, simply add the corresponding `ggplot2` (or `ggsci`) scale, which overrides the `ospsuite.plots` colors:

```{r ggplot-scale-fill-grey, fig.alt="Tile plot using ggplot2's built-in scale_fill_grey() function. The plot shows a 3x3 grid of tiles demonstrating how to override the ospsuite.plots colors with a standard ggplot2 gray scale for an individual plot."}
ggplot() +
  geom_tile(aes(x = rep(seq(1, 3), 3), y = rep(seq(1, 3), each = 3), fill = as.factor(seq(1, 9)))) +
  theme(legend.position = "none", axis.title = element_blank()) +
  scale_fill_grey() +
  labs(title = "scale_fill_grey", tag = "C")
```

```{r ggsci-lancet-colors, fig.alt="Tile plot using the ggsci Lancet journal color palette. The plot shows a 3x3 grid of tiles demonstrating how to apply a professional journal color scheme inspired by The Lancet to an individual plot with ggsci::scale_fill_lancet()."}
ggplot() +
  geom_tile(aes(x = rep(seq(1, 3), 3), y = rep(seq(1, 3), each = 3), fill = as.factor(seq(1, 9)))) +
  theme(legend.position = "none", axis.title = element_blank()) +
  ggsci::scale_fill_lancet() +
  labs(title = "ggsci::scale_fill_lancet", tag = "D")
```

For styling many unrelated plots at once, `setDefaultColorMapDistinct()` and `resetDefaultColorMapDistinct()` set or restore the discrete color palette globally via `ggplot2.discrete.*` options. They are kept for that global-styling use case; for `ospsuite.plots` plots and your own plots, prefer the per-plot `scale_colour_osp()` / `scale_fill_osp()` shown above.

## 2.4 Default Shapes

All point layers produced by `ospsuite.plots` use `geom_point_osp()`, which draws from the OSP shape palette defined in `ospShapeNames`. When you build your own plots and want them to combine consistently with `ospsuite.plots` outputs, prefer `geom_point_osp()` over the raw `ggplot2::geom_point()`.

Shape ordering is controlled per plot, not via a global option:

- `geom_point_osp()` paired with `scale_shape_osp()` assigns shapes automatically from `ospShapeNames`.
- `scale_shape_osp_manual(values = c(level1 = "circle", level2 = "diamond", ...))` sets an explicit mapping.
- `scale_shape_osp_identity()` is the right choice when the data already contains shape names.

Raw `ggplot2::geom_point()` keeps using ggplot2's native pch shapes and is unaffected.

## 2.5 Default Options

`getDefaultOptions()` returns a list of options used in this package. The `ospsuite.plots` plot functions read these options at call time (falling back to these defaults when an option is unset), so you can override them either per call or by setting the option for the session.

```{r, eval=FALSE}
# inspect the default options
ospsuite.plots::getDefaultOptions()
```

The names of all options defined by this package start with the package name `ospsuite.plots` as a prefix and a suffix. The suffixes are listed in the enumeration `OptionKeys`. There are two helper functions (`setOspsuite.plots.option` and `getOspsuite.plots.option`) to set and get these options.

## 2.5.1 Options to Customize Watermark

All plots in this packages are created with the function `ggplotWithWatermark()` instead of `ggplot()`. This function creates a `ggplot` with a customized print function which adds a watermark. 

The watermark is **enabled by default** — plots produced without any configuration will already carry the watermark.
Setting the option to `NULL` is treated the same as using the default, so the watermark will still appear.

No opt-in is required; you only need to interact with these options if you want to customize or remove the watermark.
To override this default in your `.Rprofile` (e.g., to disable watermarks globally):

```r
# Disable watermarks globally
options(ospsuite.plots.watermarkEnabled = FALSE)
```

You can edit your `.Rprofile` with `usethis::edit_r_profile()`.

**Attention!** If you combine plots e.g. with `cowplot:plot_grid` the default print function without watermark is called. In this case you have to add the watermark with the function `addWatermark(plotObject)` before the print.

The watermark can be customized by these options:

- Switch the watermark on and off (option key = `watermarkEnabled`, default = `TRUE`)
- Define the label (option key = `watermarkLabel`, default = "preliminary analysis")
- Customize format (option key = `watermarkFormat`, default = `list(x = 0.5, y = 0.5, color = "lightgrey", angle = 30, fontsize = 12, alpha = 0.7)`)

### Examples to Customize Watermark

- A: Change format of watermark
- B: Disable watermark
- C: Reset to default

```{r customize-watermark, fig.alt="Plot demonstrating customized watermark formatting. The plot shows a blank coordinate system with a red watermark rotated 90 degrees, positioned at specific coordinates with increased font size and reduced transparency, labeled 'NEW' instead of the default text."}
# Change format and label of watermark
setOspsuite.plots.option(optionKey = OptionKeys$watermarkFormat, value = list(x = 0.2, y = 0.6, color = "red", angle = 90, fontsize = 24, alpha = 0.2))
setOspsuite.plots.option(optionKey = OptionKeys$watermarkLabel, value = "NEW")

# Initialize plot
ggplotWithWatermark() + labs(title = "Changed Watermark", tag = "A")
```

```{r disable-watermark, fig.alt="Plot demonstrating disabled watermark functionality. The plot shows a clean blank coordinate system with no watermark text overlay, illustrating how the watermarkEnabled option can be set to FALSE for clean publication-ready plots."}
# Disable watermark
setOspsuite.plots.option(optionKey = OptionKeys$watermarkEnabled, value = FALSE)

# Initialize plot
ggplotWithWatermark() + labs(title = "No Watermark", tag = "B")
```

```{r reset-watermark-default, fig.alt="Plot demonstrating reset watermark to default settings. The plot shows a blank coordinate system with the default watermark restored - gray text reading 'preliminary analysis' at 30-degree angle in the center with standard transparency and font size."}
# Reset to default
setOspsuite.plots.option(optionKey = OptionKeys$watermarkFormat, value = NULL)
setOspsuite.plots.option(optionKey = OptionKeys$watermarkLabel, value = NULL)
setOspsuite.plots.option(optionKey = OptionKeys$watermarkEnabled, value = TRUE)

# Initialize plot
ggplotWithWatermark() + labs(title = "Default Watermark", tag = "C")
```

## 2.5.2 Options to Set the Defaults for Geom Layer Attributes

All plot functions have input variables `geom*Attributes` which are passed as variables to the corresponding ggplot layer. The defaults of the input variables can be set by options.

```{r options_defaultgeoms, echo=FALSE}
df <- rbind(
  data.frame(defaultOption = "Line", inputVariable = c("geomLineAttributes"), functions = c("plotTimeProfile()")),
  data.frame(defaultOption = "Ribbon", inputVariable = "geomRibbonAttributes", functions = c("plotTimeProfile()")),
  data.frame(defaultOption = "Point", inputVariable = "geomPointAttributes", functions = c("plotTimeProfile()", "plotResVsCov()", "plotRatioVsCov()", "plotPredVsObs()")),
  data.frame(defaultOption = "Errorbar", inputVariable = "geomErrorbarAttributes", functions = c("plotTimeProfile()", "plotResVsCov()", "plotRatioVsCov()", "plotPredVsObs()")),
  data.frame(defaultOption = "LLOQ", inputVariable = "geomLLOQAttributes", functions = c("plotTimeProfile()", "plotResVsCov()", "plotRatioVsCov()", "plotPredVsObs()")),
  data.frame(defaultOption = "Hist", inputVariable = "geomHistAttributes", functions = c("plotHistogram()")),
  data.frame(defaultOption = "Boxplot", inputVariable = c("geomBoxplotAttributes", "geomPointAttributes"), functions = c("plotBoxWhisker()", "plotBoxWhisker()")),
  data.frame(defaultOption = "ComparisonLine", inputVariable = c("geomComparisonLineAttributes"), functions = c("plotResVsCov()", "plotRatioVsCov()", "plotPredVsObs()")),
  data.frame(defaultOption = "GuestLine", inputVariable = c("geomGuestLineAttributes"), functions = c("plotResVsCov()", "plotRatioVsCov()", "plotPredVsObs()"))
) |>
  data.table::setDT() |>
  (\(x) x[, .(inputVariable = paste(inputVariable, collapse = ", ")), by = c("defaultOption", "functions")])() |>
  tidyr::pivot_wider(names_from = defaultOption, values_from = inputVariable, values_fill = " ")

knitr::kable(df, align = "c", caption = "Usage of options for geom attributes")
```

With default options:

- `LineAttributes = list()`
- `Ribbon = list(color = NA)`
- `PointAttributes = list()`
- `ErrorbarAttributes = list(width = 0)`
- `LLOQAttributes = list()`
- `ComparisonLineAttributes = list(linetype = "dashed")`
- `GuestLineAttributes = list(linetype = "dashed")`
- `BoxplotAttributes = list(position = position_dodge(width = 1), color = "black")`
- `HistAttributes = list(bins = 10, position = ggplot2::position_nudge())`

## 2.5.3 Options to Set Defaults for Aesthetics

Options to set the face alpha of ribbons, filled points and options to set the filled points for values below and above LLOQ.

```{r default-aesthetics, eval=FALSE}
# Default alpha = 0.5
getOspsuite.plots.option(optionKey = "alpha")

# Alpha of LLOQ values
c("TRUE" = 0.3, "FALSE" = 1)
getOspsuite.plots.option(optionKey = "lloqAlphaVector")

# Linetype LLOQ comparison lines
"dashed"
getOspsuite.plots.option(optionKey = "lloqLineType")
```

## 2.5.4 Options for Percentiles

There are two percentile-related options controlling default behavior:

- `percentiles`: A numeric vector of five quantiles defining the lower whisker, lower box, median, upper box, and upper whisker used by `plotBoxWhisker()` (default = `c(0.05, 0.25, 0.5, 0.75, 0.95)`).
- `defaultPercentiles`: A numeric vector of three quantiles used as the default for `plotRangeDistribution()` (default = `c(0.05, 0.5, 0.95)`).

```{r default-percentiles, eval=FALSE}
# Get current box-whisker percentiles (5 values)
getOspsuite.plots.option(optionKey = OptionKeys$percentiles)

# Change box-whisker percentiles globally
setOspsuite.plots.option(optionKey = OptionKeys$percentiles, value = c(0.1, 0.25, 0.5, 0.75, 0.9))

# Get current range-plot percentiles (3 values)
getOspsuite.plots.option(optionKey = OptionKeys$defaultPercentiles)

# Change range-plot default percentiles globally
setOspsuite.plots.option(optionKey = OptionKeys$defaultPercentiles, value = c(0.1, 0.5, 0.9))
```

## 2.5.5 Options to Define Export Format

There are options to define the export format using the function `exportPlot` (see details below):

- `exportWidth`: Width of the exported file
- `exportUnits`: Units for width and height
- `exportDevice`: Device for plot export
- `exportDpi`: Plot resolution

The latter options are used directly as input for `ggplot2::ggsave`, so check the help for available values.

# 3. Plot Functions

All plot functions listed below call internally the function `initializePlot()`. This function constructs labels from the metadata and adds a watermark layer. It can also be used to create a customized ggplot.

## 3.1 `plotTimeProfile()`

[Time Profile Plots](plot-time-profile.html)

## 3.2 `plotBoxWhisker()`

[Box Whisker Plots](box-whisker-vignette.html)

## 3.3 `plotHistogram()`

[Histogram Plots](histogram.html)

## 3.4 `plotPredVsObs()`

[Goodness of Fit](Goodness_of_fit.html)

## 3.5 `plotResVsCov()`

[Goodness of Fit](Goodness_of_fit.html)

## 3.6 `plotRatioVsCov()`

[Goodness of Fit](Goodness_of_fit.html)

## 3.7 `plotQQ()`

[Goodness of Fit](Goodness_of_fit.html)

# 4. Additional Aesthetics

This package provides some additional aesthetics.

For more details, see the examples within the vignettes for the respective functions.

- `groupby`: Shortcut to use different aesthetics to group. All functions where this aesthetic is used have also a variable `groupAesthetics`. The mapping `groupby` is copied to all aesthetics listed within this variable and to the aesthetic `group`.
- `lloq`: Mapped to a column with values indicating the lower limit of quantification, adds horizontal (or vertical) lines to the plot. All observed values below the "lloq" are plotted with a lighter alpha. As values are compared row by row, it is possible to have more than one LLOQ.
- `error`: Mapped to a column with additive error (e.g., standard deviation); error bars are plotted. This is a shortcut to map `ymin` and `ymax` directly (`ymin = y - error` and `ymax = y + error`). Additionally, if `yScale` is set, `ymin` values below 0 are set to `y`.
- `error_relative`: Mapped to a column with relative error (e.g., geometric standard deviation); error bars are plotted. This is a shortcut to map `ymin` and `ymax` directly (`ymin = y / error_relative` and `ymax = y * error_relative`).
- `y2axis`: Creates a plot with 2 y axes. It is used to map to a column with a logical value. Values where this column has a TRUE entry will be displayed with a secondary axis.
- `mdv`: Mapped to a logical column. Rows where this column has entries set to TRUE are not plotted (MDV = missing data value, taken from NONMEM notation).
- `observed` / `predicted`: For the function `plotPredVsObs()`, `observed` is mapped to `x` and `predicted` is mapped to `y`.

Residuals and ratios must be pre-calculated before passing them to the plotting functions. If you are using the `{ospsuite}` package, use `ospsuite::addResidualColumn()` to add a residual column to your data.

See `vignette("Goodness of Fit", package = "ospsuite.plots")` for examples.

```{r additional-aesthetic-table, echo=FALSE}
df <- rbind(
  data.frame(aesthetic = "`groupby`", functions = c("plotTimeProfile()", "plotHistogram()", "plotPredVsObs()", "plotResVsCov()", "plotRatioVsCov()", "plotQQ()")),
  data.frame(aesthetic = "`lloq`", functions = c("plotTimeProfile()", "plotPredVsObs()")),
  data.frame(aesthetic = "`error`", functions = c("plotTimeProfile()", "plotRatioVsCov()", "plotPredVsObs()", "plotResVsCov()")),
  data.frame(aesthetic = "`error_relative`", functions = c("plotTimeProfile()", "plotRatioVsCov()", "plotPredVsObs()", "plotResVsCov()")),
  data.frame(aesthetic = "`y2axis`", functions = "plotTimeProfile()"),
  data.frame(aesthetic = "`mdv`", functions = c("plotTimeProfile()", "plotBoxWhisker()", "plotHistogram()", "plotPredVsObs()", "plotResVsCov()", "plotRatioVsCov()", "plotQQ()")),
  data.frame(aesthetic = "`observed` / `predicted`", functions = c("plotPredVsObs()"))
) |>
  dplyr::mutate(used = "X") |>
  tidyr::pivot_wider(names_from = aesthetic, values_from = used, values_fill = " ")

knitr::kable(df, align = "c", caption = "Applicability of additional aesthetics in functions")
```

# 5. Plot Export

This section demonstrates how to use the `exportPlot` function to save `ggplot` objects to files, adjusting the width and height of the exported plots as necessary. This function is part of the `ospsuite.plots` package and simplifies the process of exporting plots for various purposes, such as publication or presentation.

## 5.1 Basic Usage

To export a plot, you need a ggplot object. Here's a basic example:

Create a simple ggplot object:

```{r export-object, eval=FALSE}
plotObject <- ospsuite.plots::plotHistogram(data = testData, mapping = aes(x = Age))
```

Exporting the Plot: Using the `exportPlot` function, you can easily save this plot to a file:

Replace "path/to/save" with the actual directory path where you want to save the plot, and adjust the width and height parameters as needed.

```{r export-basic, eval=FALSE}
exportPlot(plotObject = plotObject, filepath = "path/to/save", filename = "myplot", width = 10, height = 8)
```

## 5.2 Advanced Usage

### Adjusting Plot Dimensions Based on Content

The `exportPlot` function can automatically adjust the plot dimensions based on its content, such as the presence of a legend or the aspect ratio.

Assuming `plotObject` is your ggplot object:

```{r export-advanced, eval=FALSE}
exportPlot(plotObject = plotObject, filepath = "path/to/save", filename = "adjusted_plot.png")
```

In this case, you don't need to specify the width and height explicitly; the function calculates them for you. The default width value saved in the `ospsuite.plots` option `exportWidth` is used. If an aspect ratio is defined in the theme of the plot, height will be adjusted accordingly; otherwise, the function exports square figures.

# 6. Shapes

## 6.1 Default Shapes

```{r default-shapes, out.width = "90%", fig.asp = 0.7, fig.width = 8, fig.alt="Chart displaying all default OSP shape types available in ospsuite.plots. Shows various point shapes arranged in a grid with labels, demonstrating the visual appearance of each shape option."}
shapes <- data.frame(
  shapeNames = ospShapeNames,
  x = rep(1:6, length.out = length(ospShapeNames)),
  y = rep(1:4, each = 6, length.out = length(ospShapeNames))
)

ggplot(shapes, aes(x, y, shape = shapeNames)) +
  geom_point_osp(size = 5, color = "blue", fill = "red") +
  scale_shape_osp_identity() +
  geom_text(aes(label = shapeNames), nudge_y = -0.3, size = 3) +
  theme_void()
```

## 6.2 Using OSP Shapes in Plot Functions

OSP shapes are applied per plot in all `ospsuite.plots` functions. You can also apply them directly to your own plots with `scale_shape_osp()`.

```{r osp-shapes-demo, fig.alt="Scatter plot demonstrating OSP shapes in ospsuite.plots. Shows four data points representing different species using distinct OSP shapes with each species having a distinct symbol and color."}
dt <- data.frame(x = c(1, 2, 1, 2), y = c(1, 1, 2, 2), species = c("dog", "cat", "mouse", "rat"))

plotObject <- plotYVsX(data = dt, mapping = aes(x = x, y = y, groupby = species), xScale = "linear", xScaleArgs = list(limits = c(0.5, 2.5)), yScale = "linear", yScaleArgs = list(limits = c(0.5, 2.5)))

plot(plotObject)
```

```


