---
title: "Time Profile Plots"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Time Profile Plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  dpi = 300,
  fig.show = "hold",
  fig.align = "center",
  fig.width = 7,
  out.width = "80%",
  comment = "#>"
)
```


# 1. Introduction

The following vignette aims at documenting and illustrating workflows for producing Time Profile plots using the `ospsuite.plots`-library.

Time profile plots are very often used to compare observed and simulated data along time.
In such plots, observed data are usually plotted as scatter points with errorbars showing population range or confidence intervals; while simulated data are usually plotted using lines with shaded ribbons showing population range or confidence intervals.

>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

Basic documentation of the function can be found using: `?plotTimeProfile`.
The output of the function is a `ggplot` object.

## 1.1 Setup

For this vignette we will use the libraries {ospsuite.plots} and {tidyr}.
We use the default settings of {ospsuite.plots} (see vignette("ospsuite.plots", package = "ospsuite.plots")) but adjust the legend position

```{r setup,warning=FALSE,message=FALSE}
library(ospsuite.plots)
library(tidyr)
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

# setDefaults
oldDefaults <- ospsuite.plots::setDefaults()


# Place default legend position above the plot for prettier time profile plots
theme_update(legend.position = "top")
theme_update(legend.direction = "vertical")
theme_update(legend.box = "horizontal")
theme_update(legend.title = element_blank())
```




## 1.2 Example data

This vignette uses random generated example data sets provided by the package:
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

### 1.2.1 simulated and observed data

In the example below theses data sets are used:

* `simData1` and `obsData1` set with an 'exponential decay'  
* `simData2` and `obsData2` set with a 'first order absorption with exponential decay'  
* `simData` and `obsData`  combination of `simData1` and `simData2` / `obsData1` and `obsData2`


```{r dataset1}
simData1 <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet1") %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::select(c("time", "values", "minValues", "maxValues", "caption"))


simData2 <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet2") %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::select(c("time", "values", "minValues", "maxValues", "caption"))


simData <- rbind(
  simData1,
  simData2
)
<<<<<<< HEAD

obsData1 <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet1") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "sd", "maxValues", "minValues", "caption"))


obsData2 <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet2") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "sd", "maxValues", "minValues", "caption"))

obsData <- rbind(
  obsData1,
  obsData2
)



knitr::kable(
  head(simData),
  digits = 3,
  caption =
    "first rows of example data simData"
)
```


* `simDataLloq` and `obsDataLloq`: data set with a column which defines  lloq


```{r dataset2}
simDataLloq <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == c("DataSet3")) %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::filter(dimension == "concentration") %>%
  dplyr::select(c("time", "values", "caption"))


obsDataLloq <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet3") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::filter(dimension == "concentration") %>%
  dplyr::select(c("time", "values", "caption", "lloq", "error_relative"))
```


* `simData2Dimension`: dataset where the "values" column has mixed dimensions: "concentration" and "fraction", 

```{r dataset3}
simData2Dimension <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet3") %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::select(c("time", "values", "dimension", "caption"))



obsData2Dimension <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet3") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "dimension", "caption", "lloq", "error_relative"))
```

* `obsDataGender` observed data set with gender information , and `simDataGender` a mean model presentation

```{r dataset4}
simDataGender <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet4") %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::select(c("time", "values", "caption"))

obsDataGender <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet4") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "caption", "gender"))
```


### 1.2.2 MetaData

Metadata is a list, which contains dimension and unit information for dataset-columns.
If available axis labels are set by this information.
If for the x axis a time Unit can be identified, 
breaks are set according to this unit. (see `?updateScaleArgumentsForTimeUnit`)


```{r meta-data}
metaData <- attr(exampleDataTimeProfile, "metaData")



knitr::kable(metaData2DataFrame(metaData), digits = 2, caption = "list of meta data")
```





# 2 Examples

The following sections show *how to plot a Time Profile* for specific scenarios

## 2.1 Plot simulated data only


### 2.1.1 Basic example with multiple simulations

Data sets mapped to data are displayed as lines:

The aesthetic `groupby` which is mapped in the example to the columns caption groups profiles by column `caption`. That means `caption` is internally mapped to all  aesthetics defined in the variable `groupAesthetics`. 
By default this are `color`, `linetype`, `shape` (only relevant for observed data) and `fill`.

```{r examples-single-mappingsimulation-plot-A}
#  A Multiple simulations

plotTimeProfile(
  data = simData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    groupby = caption
  )
)
```


### 2.1.2  Multiple simulations with confidence interval

Mapping `ymin` and `ymax` will add a ribbon to the time profile which may indicate a prediction confidence interval or a population variance. 

```{r examples-single-mappingsimulation-plot-B}
plotTimeProfile(
  data = simData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    ymin = minValues,
    ymax = maxValues,
    groupby = caption
  )
)
```



## 2.2 Plot observed data only
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

### 2.2.1  Basic example with multiple observed data sets

A data set mapped to observed data is displayed as points.


```{r examples-single-observation-A}
plotTimeProfile(
  observedData = obsData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    groupby = caption
  )
)
```


### 2.2.2 Observed data sets with confidence interval

mapping of `ymin` and `ymax`  adds errorbars


```{r examples-single-observation-B}
plotTimeProfile(
  observedData = obsData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    ymin = minValues,
    ymax = maxValues,
    groupby = caption
  )
)
```

### 2.2.3 Usage of aesthetic "error"

The data set has also a columns with the standard deviation sd.
If mapped to 'error' this variable will be used to create corresponding `ymin` and `ymax` values for the error bars $ymin = values - sd$, $ymax = values + sd$.  If `yscale` = 'log', `ymin` values below 0 are set to `y`.  
Also available is error_relative where a multiplicative error is assumed: $ymin = \frac{values}{error\_relative}$, $ymax = values*error\_relative$.



```{r examples-single-observation-C}
plotTimeProfile(
  observedData = obsData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    error = sd,
    groupby = caption
  )
)
```


### 2.2.4 Observed data with LLOQ

If `lloq` is mapped to a column with values indicating the lower limit of quantification , a horizontal line for the `lloq` values are displayed, and all values below `lloq` are plotted with decreased alpha.
As the comparison is done by row, multiple lloq values are possible.


```{r example-lloq}
plotTimeProfile(
  observedData = obsDataLloq,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    groupby = caption,
    error_relative = error_relative,
    lloq = lloq
  ),
  yscale = "log"
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette
)
```


>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

### 2.2.5 Omit data points flagged as Missing Dependent Variable (mdv)

The following code adds a new column where all values higher than `10` are flagged as `mdv`. This leads to a plot without any observed data points higher than `10` (removing the first observation)


```{r examples mdv}
# Use common variable before usinf rbind.data.frame
mdvData <- obsData
mdvData$mdv <- mdvData$values > 10

plotTimeProfile(
  observedData = mdvData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    ymin = minValues,
    ymax = maxValues,
    groupby = caption,
    mdv = mdv
  )
)
```


## 2.3 Plot simulated and observed data

By plotting simulated and observed data together in most cases you have pairs of corresponding simulated and observed data sets. This can either be done by using a common legend entry (see section 2.3.1) or by defining a mapping table, where each observed data sets is mapped to one simulated data set (see section 2.3.2).  
There may be some example with independent data sets (see section 2.3.3)


### 2.3.1 Corresponding simulated and observed datasets with common legend entry

In this example we create as first step in each data set a column with a common caption for simulated and observed data.
The new column is then mapped to aesthetic `groupby`. That leads to one common legend for observed and simulated data.

```{r combine-simulated-and-observed-sharing-legend}
# create datasets with common caption
simData <- data.frame(simData) %>%
  dplyr::mutate(captionCommon = gsub("Simulated ", "", caption))

obsData <- data.frame(obsData) %>%
  dplyr::mutate(captionCommon = gsub("Observed ", "", caption))


plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    ymin = minValues,
    ymax = maxValues,
    groupby = captionCommon
  )
)
```



### 2.3.2 Corresponding simulated and observed datasets with separate legend entries and mapping table

In this example we create as first step a mapping table, with one column `observed` and one column `simulated`.
This new table is passed to the function as input variable `mapSimulatedAndObserved`. 


```{r combine-simulated-and-separate-sharing-legend}
mapSimulatedAndObserved <- data.frame(
  simulated = unique(simData$caption),
  observed = unique(obsData$caption)
)

knitr::kable(mapSimulatedAndObserved)


plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping <- aes(
    x = time,
    y = values,
    ymin = minValues,
    ymax = maxValues,
    groupby = caption
  ),
  mapSimulatedAndObserved = mapSimulatedAndObserved
)
```


If not for all simulated data sets exists observed data sets (or vice versa), it is possible to fill the mapping table with empty strings for the missing data sets. The empty strings have to be at the end of the table.


```{r combine-simulated-and-observed-separate-legend-2}
mapSimulatedAndObserved <- data.frame(
  simulated = unique(simData$caption),
  observed = c(unique(obsData1$caption), "")
)


knitr::kable(mapSimulatedAndObserved)

plotTimeProfile(
  data = simData,
  observedData = obsData1,
  metaData = metaData,
  mapping <- aes(
    x = time,
    y = values,
    ymin = minValues,
    ymax = maxValues,
    groupby = caption
  ),
  mapSimulatedAndObserved = mapSimulatedAndObserved
)
```




### 2.3.3 Independent simulated and observed datasets

The example below shows individual observed data compared to one simulation. Here a mapping between observed and simulated doesn't make sense. 
`groupby` is used to group the observed data by color, fill and shape.
As we do not want to have an extra color for the simulated line, the line color is set by `geomLineAttributes` as attribute and not as aesthetic. The simulated line inherits the aesthetic `linetype` from `groupby`and therefore a `linetype` legend is added which contains the simulated data.

```{r combine-multiple-simulated-and-observed}
plotTimeProfile(
  data = simDataGender,
  observedData = obsDataGender,
  metaData = metaData,
  mapping <- aes(
    x = time,
    y = values,
    groupby = caption
  ),
  geomLineAttributes = list(color = "black")
) +
  theme(legend.position = "right") +
  labs(
    color = "observed data",
    shape = "observed data",
    fill = "observed data",
    linetype = "simulation"
  ) +
  theme(legend.title = element_text())
```

>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

For changing scales which ared adressed by the the grouping, the reset of scales after the simulation layers has to be taken in account. (see [Plot simulated and observed data](#s_groupAesthtics)).
It's a stepwise procedure

### 2.3.4 Multiple simulations and observed data sets without legends

To map `groupby` with an empty variable `groupAesthetics` leads to a plot without legends.

```{r examples-multiple-simulations-no-caption}
plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    groupby = caption,
  ), groupAesthetics = c()
)
```
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

plotObject <- plotTimeProfile(
  plotObject = plotObject,
  observedData = obsData,
  mapping = mapping
) +
  scale_color_manual(values = c("Observed Data 1" = "darkred", "Observed Data 2" = "darkgreen"))

>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette


### 2.3.5 Observed data with shape as gender

>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette


In this example observed data is used as simulated and as observed data, to connect the different data points with a thin line.

```{r observedData-usead-as-simulatedData-plot}
plotTimeProfile(
  data = obsDataGender,
  observedData = obsDataGender,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    groupby = caption,
    shape = gender
  ),
  geomLineAttributes = list(linetype = "solid", linewidth = 0.5)
) + theme(legend.position = "right")
```

A similar plot can produces by combining caption and gender with interaction

In the example we set the scale for the y-axis to log scale.
Per default a time profile plots starts always at 0, here the defaults were overwritten and the breaks were set manually.

```{r observedData-usead-as-simulatedData-plot-B}
plotTimeProfile(
  data = obsDataGender,
  observedData = obsDataGender,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    groupby = interaction(caption, gender)
  ),
  geomLineAttributes = list(linetype = "solid", linewidth = 0.5)
) + theme(legend.position = "right")
```



### 2.3.6 Data with Secondary axis


In this example a plot is generated with concentration on the left y-axis and Fraction on the right y-axis. As first step the variable `metaData` has to be adjusted.
For the primary y-axis the column "values" is mapped, and metaData provides for this column the dimension "Concentration". For the secondary y-axis a new entry "y2" is added.

```{r secondrayAxis_metaData}
metaData_y2 <- list(
  time = list(
    dimension = "Time",
    unit = "h"
  ),
  values = list(
    dimension = "Concentration",
    unit = "mg/l"
  ),
  y2 = list(
    dimension = "Fraction",
    unit = ""
  )
)
```


The mapping `y2axis` has to be logical. In this example it is `(dimension == "fraction")`.
For the primary y-axis (concentration) a log scale is displayed and for the secondary (Fraction) a linear scale. Limits of the secondary-axis are set to [0,1]

```{r secondrayAxis_plot}
plotTimeProfile(
  data = simData2Dimension,
  observedData = obsData2Dimension,
  mapping = aes(
    x = time,
    y = values,
    error_relative = error_relative,
    lloq = lloq,
    y2axis = (dimension == "fraction"),
    groupby = dimension,
    shape = caption
  ),
  metaData = metaData_y2,
  yscale = "log",
  yscale.args = list(limits = c(0.01, NA)),
  y2scale = "linear",
  y2scale.args = list(limits = c(0, 1)),
) +
  theme(
    axis.title.y.right = element_text(angle = 90),
    legend.position = "right",
    legend.box = "vertical"
  )
```


# 3. Plot Configuration


## 3.1 Example for changing geom attributes

The plot from section 2.3.2  was adjusted using geom Attributes:

- `geomLineAttributes = list(linetype = 'solid')` The lines used for the simulated data are set in both datasets to solid. Attention the line type for the errorbars and ribbon edges was not changed.
- `geomErrorbarAttributes = list()` The default settings for `geomErrorbarAttributes`, width = 0, was deleted. The bar caps are now visible
- `geomRibbonAttributes = list(alpha = 0.1)` the shade of the ribbons was decreased by setting the alpha to 0.1, the default values for color = NA, was omitted, so that the edges are now visible
- `geomPointAttributes = list(size = 7)` The size of the symbols are increased


```{r changing_geom_attributes}
mapSimulatedAndObserved <- data.frame(
  simulated = unique(simData$caption),
  observed = rev(unique(obsData$caption))
)


plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    ymin = minValues,
    ymax = maxValues,
    groupby = caption
  ),
  geomLineAttributes = list(linetype = "solid"),
  geomErrorbarAttributes = list(width = 3),
  geomRibbonAttributes = list(alpha = 0.1),
  geomPointAttributes = list(size = 7),
  mapSimulatedAndObserved = mapSimulatedAndObserved
)
```



## 3.2 Example for changing color scales


## 3.2.1  Without mapping table

For plots showing only simulated or only observed or plots with a common legend (see section 2.3.1) the colors are changed using {ggplot2} functions like `scale_color_manual`


Below the plot from section 2.3.1 is repeated and 

```{r colorscaling}
# create datasets with common caption
simData <- data.frame(simData) %>%
  dplyr::mutate(captionCommon = gsub("Simulated ", "", caption))

obsData <- data.frame(obsData) %>%
  dplyr::mutate(captionCommon = gsub("Observed ", "", caption))


plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    ymin = minValues,
    ymax = maxValues,
    groupby = captionCommon
  )
) +
  scale_color_manual(values = c("Data 1" = "darkred", "Data 2" = "darkgreen")) +
  scale_fill_manual(values = c("Data 1" = "red", "Data 2" = "green"))
```

## 3.2.2  With mapping table

It is possible to add columns with aesthetics to the table used to map simulated and observed data.
The column header must correspondent to one of the aesthetics defined in `groupgroupAesthetics`.

In the example below this is done for 'color' and 'fill':

```{r colorscaling-with-mappingtable}
# Define Data Mappings

mapSimulatedAndObserved <- data.frame(
  simulated = unique(simData$caption),
  observed = unique(obsData$caption),
  color = c("darkred", "darkgreen"),
  fill = c("red", "green")
)

knitr::kable(mapSimulatedAndObserved)


plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    ymin = minValues,
    ymax = maxValues,
    groupby = caption
  ),
  mapSimulatedAndObserved = mapSimulatedAndObserved
)
```


Changing the scales when using the observed simulation mapping table  can also done by adding scale manually, but it is a bit more complicated. If `mapSimulatedAndObserved` is not null, a reset of all relevant scales is done before plotting the observed data. The scales for the simulated data has to be done before this reset. You have to call `plotTimeProfile` two times:

- call `plotTimeProfile()` for simulated data only
- set scales for simulated data
- call `plotTimeProfile()` for observed data only with the simulated plot as input  `plotObject`
- set scales for observed data



```{r example_changing_colorscales}
mapSimulatedAndObserved <- data.frame(
  simulated = unique(simData$caption),
  observed = rev(unique(obsData$caption))
)

# Define Data Mappings
mapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  groupby = caption
)


plotObject <- plotTimeProfile(
  data = simData,
  metaData = metaData,
  mapping = mapping,
  mapSimulatedAndObserved = mapSimulatedAndObserved
) +
  scale_color_manual(values = c("Simulated Data 1" = "darkred", "Simulated Data 2" = "darkgreen")) +
  scale_fill_manual(values = c("Simulated Data 1" = "red", "Simulated Data 2" = "green"))

plotObject <- plotTimeProfile(
  plotObject = plotObject,
  observedData = obsData,
  mapping = mapping,
  mapSimulatedAndObserved = mapSimulatedAndObserved
) +
  scale_color_manual(values = c("Observed Data 1" = "darkred", "Observed Data 2" = "darkgreen"))

plot(plotObject)
```





## 3.3 Example for adjusting x and y scale

In the example we set the scale for the y-axis to log scale.
Per default a time profile plots starts always at 0, here the defaults were overwritten and the breaks were set manually.

```{r example_adjsuting_xy_scale}
plotTimeProfile(
  data = simData %>%
    dplyr::filter(values > 0),
  metaData = metaData,
  mapping = aes(
    x = time + 24,
    y = values,
    groupby = caption
  ),
  yscale = "log",
  xscale.args = list(
    limits = c(24, 48),
    breaks = seq(24, 48, 3)
  )
)
```



## 3.4 Adjust Time unit

The breaks of time axis are set according to the units provided by the variable `metaData`.

Below we show the same plot with 4 different time units:

```{r example_timeUnit,fig.height=7}
plotlist <- list()

for (unit in c("h", "day(s)", "weeks(s)", "month(s)")) {
  metaData$time$unit <- unit

  plotlist[[unit]] <- plotTimeProfile(
    data = simData,
    metaData = metaData,
    mapping = aes(
      x = time,
      y = values,
      groupby = caption
    )
  ) +
    labs(
      title = paste("Time unit:", unit)
    ) +
    theme(legend.position = "none")
}

cowplot::plot_grid(plotlist = plotlist, labels = "AUTO")
```

```{r cleanup,echo=FALSE}
resetDefaults(oldDefaults)
```

