---
title: "Time Profile Plots"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Time Profile Plots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.show = "hold",
  comment = "#>"
)
```

```{r setup,warning=FALSE,message=FALSE}
library(ospsuite.plots)
library(tidyr)

# setDefaults
ospsuite.plots::setDefaults()


# Place default legend position above the plot for prettier time profile plots
theme_update(legend.position = "top")
theme_update(legend.title = element_blank())
```


# 1. Introduction

The following vignette aims at documenting and illustrating workflows for producing Time Profile plots using the `ospsuite.plots`-Library.

Time profile plots are very often used to compare observed and simulated data along time.
In such plots, observed data are usually plotted as scatter points with errorbars showing population range or confidence intervals; while simulated data are usually plotted using lines with shaded ribbons showing population range or confidence intervals.

<<<<<<< HEAD
# 1.1 Inputs for plotTimeProfile
=======
# 1.1 Inputs for function `plotTimeProfile`
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette


inputs for data mapping and descriptions
* `data`  data.frame with simulated data
* `observedData` data.frame with observed data
* `mapping`  Default list of aesthetic mappings to use for plot
* `observedMapping` Default list of aesthetic mappings to use for observed data
* `metaData` A named list of information about `data` such as the `dimension` and `unit` of its variables.

<<<<<<< HEAD
In this vignette simulated and observed data have corresponding column names. We can use a common mapping for both datasets. If this is not the case the variabale `observedMapping` has to be used for the observed data.

there are some mappings additional mappings to the ones provided by ggplot: 
  * lloq see example [2.4 Observed data with LLOQ]
  * mdv see example [2.2.5. Missing Dependent Variable (mdv)]
  * error, error_relativ see example [2.2.2. Single simulation with confidence interval]
  * y2axis see example [Data with Secondary axis]
=======
In this vignette simulated and observed data have corresponding column names. We can use a common mapping for both data sets. If this is not the case the variable `observedMapping` has to be used for the observed data.

there are some mappings additional mappings to the ones provided by ggplot: 
  * `lloq` see example [2.4 Observed data with LLOQ]
  * `mdv` see example [2.2.5. Missing Dependent Variable (mdv)]
  * `error`, `error_relative` see example [2.2.2. Single simulation with confidence interval]
  * `y2axis` see example [Data with Secondary axis]
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette


* `plotObject` An optional `ggplot` object on which to add the plot layers

inputs for plot configuration:

* `geomLineAttributes` A `list` with arguments which are passed on to the ggplot2::geom_line
* `geomRibbonAttributes` A `list` with arguments which are passed on to the ggplot2::geom_ribbon
* `geomPointAttributes` A `list` with arguments which are passed on to the ggplot2::geom_point
* `geomErrorbarAttributes` A `list` with arguments which are passed on to the ggplot2::geom_errorbar
* `geomLLOQAttributes` A `list` with arguments which are passed on to the ggplot2::geom_hline

see example [3.1 Example for changing geom attributes]

     
<<<<<<< HEAD
* groupAesthetics vector of aesthetics, which are used for columns mapped with group,
=======
* `groupAesthetics` vector of aesthetics, which are used for columns mapped with group,
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette
     use of group aesthetics triggers second axis after simulation layers
     
see example [2.3. Plot simulated and observed data]

inputs to scale x and y axis:

<<<<<<< HEAD
* xscale either 'linear' then ggplot2::scale_x_continuous() or 'log' then ggplot2::scale_x_log10() is used
* xscale.args list of arguments passed to ggplot2::scale_x_continuous() or ggplot2::scale_x_log10()
* yscale either 'linear' then ggplot2::scale_y_continuous() or 'log' then ggplot2::scale_y_log10() is used
* yscale.args list of arguments passed to ggplot2::scale_y_continuous() or ggplot2::scale_y_log10()
* y2scale either 'linear' the secondary axis is displayed linear, or 'log' secondary axis is displayed with log scale
* y2scale.args list of arguments passed to ggplot2::sec_axis(), trans, break are set by code
=======
* `xscale` either 'linear' then ggplot2::scale_x_continuous() or 'log' then ggplot2::scale_x_log10() is used
* `xscale.args` list of arguments passed to ggplot2::scale_x_continuous() or ggplot2::scale_x_log10()
* `yscale` either 'linear' then ggplot2::scale_y_continuous() or 'log' then ggplot2::scale_y_log10() is used
* `yscale.args` list of arguments passed to ggplot2::scale_y_continuous() or ggplot2::scale_y_log10()
* `y2scale` either 'linear' the secondary axis is displayed linear, or 'log' secondary axis is displayed with log scale
* `y2scale.args` list of arguments passed to ggplot2::sec_axis(), trans, break are set by code
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

see example [3.3 Example for adjusting x and y scale]


# 1.2 Example data

This vignette uses random generated example data provided by the package, which is loaded at the beginning of each section


The following sections show *how to plot a Time Profile* for specific scenarios

## 2.1. Plot simulated data only


```{r simulation-data}
simData1 <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet1") %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))

simData2 <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet2") %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))

knitr::kable(head(simData1), digits = 3, caption = "first rows of example data simData1")

metaData <- attr(exampleDataTimeProfile, "metaData")
knitr::kable(metaData2DataFrame(metaData), caption = "list of meta data:")
```
<<<<<<< HEAD

### 2.1.1 Single simulation


```{r examples-single-mappingsimulation-plot}
# Define Data Mapping
mapping <- aes(
  x = time,
  y = values,
  group = caption
)

plotTimeProfile(
  data = simData1,
  metaData = metaData,
  mapping = mapping,
)
```

### 2.1.2. Single simulation with confidence interval

```{r examples-single-simulation-with-confidence-interval}
# Define Data Mapping
simDataMapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption
)

plotTimeProfile(
  data = simData1,
  metaData = metaData,
  mapping = simDataMapping
)
```





### 2.1.3. Multiple simulations

Using the aesthetic group in timeprofiles will automatically discriminate grouped profiles by 
color, linetype, shape and fill.


```{r examples-multiple-simulations}
# Define Data Mapping

simDataMapping <- aes(
  x = time,
  y = values,
  group = caption
)

plotTimeProfile(
  data = rbind.data.frame(
    simData1,
    simData2
  ),
  metaData = metaData,
  mapping = simDataMapping
)
```
### 2.1.4. Multiple simulations with confidence interval

```{r examples-multiple-simulations-with-confidence-interval}
# Define Data Mapping
# ymin and ymax define the CI
simDataMapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption
)

plotObject <- plotTimeProfile(
  data = rbind.data.frame(
    simData2,
    simData1
  ),
  metaData = metaData,
  mapping = simDataMapping
)

plot(plotObject)
```



## 2.2. Plot observed data only



```{r observed-data}
obsData1 <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet1") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "sd", "caption"))

obsData2 <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet2") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))

obsData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID %in% c("DataSet1", "DataSet2")) %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))


knitr::kable(head(obsData1), digits = 3, caption = "first rows of example data obsData1")
knitr::kable(head(obsData2), digits = 3, caption = "first rows of example data obsData2")

metaData <- attr(exampleDataTimeProfile, "metaData")
knitr::kable(metaData2DataFrame(metaData), caption = "list of meta data:")
```


### 2.2.1. Single observed data set

```{r examples-single-observation}
# Define Data Mapping
obsAes <- aes(
  x = time,
  y = values,
  group = caption
)

plotTimeProfile(
  observedData = obsData1,
  metaData = metaData,
  observedMapping = obsAes
)
```



### 2.2.2. Single observed data set with confidence interval

For this  data set, the uncertainty/error was defined as standard deviation using the variable `"sd"`.
If mapped to 'error' this variable will be used to create corresponding `ymin` and `ymax` values for the error bars $ymin = values - sd$, $ymax = values + sd$. 
Also available is error_relative where a multiplikative error is assumed: $ymin = values*error_relative$, $ymax = values*error\_relative$

**Tips**

- For data that include errorbars only partially, `NA` values cannot be used and should be replaced by `0` for error variable or `y` for `ymin` and `ymax` variables. 
- For logarithmic plots, negative `ymin` values of errorbars should be replaced by `y` beforehand in order to still see the `ymax` range of the errorbars.

if the mapping `error` is used and the variable `yscale` is set to 'log', all 'ymin' values < 0 are set to 'y'



```{r examples-single-observation-with-confidence-interval-as-error}
# Define Data Mapping
obsDataMapping1 <- aes(
  x = time,
  y = values,
  error = sd,
  group = caption
)

plotTimeProfile(
  observedData = obsData1,
  metaData = metaData,
  mapping = obsDataMapping1
)
```


For a second dataset, the uncertainty/error was defined as `"minValues"` and `"maxValues"`, they are input directly as `ymin` and `ymax` and will be taken as is.

```{r examples-single-observation-with-confidence-interval-as-ymin-ymax}
# Define Data Mapping
obsDataMapping2 <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
=======

### 2.1.1 Single simulation


```{r examples-single-mappingsimulation-plot}
# Define Data Mapping
mapping <- aes(
  x = time,
  y = values,
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette
  group = caption
)

plotTimeProfile(
<<<<<<< HEAD
  observedData = obsData2,
  metaData = metaData,
  mapping = obsDataMapping2
)
```

### 2.2.3. Multiple observed data sets

```{r examples-multiple-observations}
# Define Data Mapping
obsDataMapping <- aes(
  x = time,
  y = values,
=======
  data = simData1,
  metaData = metaData,
  mapping = mapping,
)
```

### 2.1.2. Single simulation with confidence interval

```{r examples-single-simulation-with-confidence-interval}
# Define Data Mapping
simDataMapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette
  group = caption
)

plotTimeProfile(
<<<<<<< HEAD
  observedData = obsData,
  metaData = metaData,
  mapping = obsDataMapping
)
```

### 2.2.4. Multiple observed data sets with confidence interval

In this example, the observed data were not using the same way of displaying errors.
Before plotting, the data needs to be merged in a consistent way to let the mapping know which variable(s) to use.
=======
  data = simData1,
  metaData = metaData,
  mapping = simDataMapping
)
```


>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

Below, `ymin` and `ymax` were consequently calculated for `obsData1` before merging.

```{r examples-multiple-observations-with-confidence-interval}
# Use common variable before use rbind.data.frame
obsData1$minValues <- obsData1$values - obsData1$sd
obsData1$maxValues <- obsData1$values + obsData1$sd
obsData2$sd <- 0

<<<<<<< HEAD
# Define Data Mapping
obsDataMapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption
)

plotTimeProfile(
  observedData = rbind.data.frame(
    obsData1,
    obsData2
  ),
  metaData = metaData,
  mapping = obsDataMapping
)
```


### 2.2.5. Missing Dependent Variable (mdv)

The following code flags all values higher than `10` as `mdv`, leading to a plot without any observed data points higher than `10` (removing the first observation)


```{r examples mdv}
# Use common variable before usinf rbind.data.frame
mdvData <- obsData1
mdvData$mdv <- mdvData$values > 10

# Define Data Mapping
obsDataMapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption,
  mdv = mdv
)

plotTimeProfile(
  observedData = mdvData,
  metaData = metaData,
  mapping = obsDataMapping
)
```


## 2.3. Plot simulated and observed data
=======
### 2.1.3. Multiple simulations

Using the aesthetic group in timeprofiles will automatically discriminate grouped profiles by 
color, linetype, shape and fill.


```{r examples-multiple-simulations}
# Define Data Mapping

simDataMapping <- aes(
  x = time,
  y = values,
  group = caption
)

plotTimeProfile(
  data = rbind.data.frame(
    simData1,
    simData2
  ),
  metaData = metaData,
  mapping = simDataMapping
)
```
### 2.1.4. Multiple simulations with confidence interval

```{r examples-multiple-simulations-with-confidence-interval}
# Define Data Mapping
# ymin and ymax define the CI
simDataMapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption
)

plotObject <- plotTimeProfile(
  data = rbind.data.frame(
    simData2,
    simData1
  ),
  metaData = metaData,
  mapping = simDataMapping
)

plot(plotObject)
```



## 2.2. Plot observed data only
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette


```{r observed-and-simulated-data}
simData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID %in% c("DataSet1", "DataSet2")) %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))

<<<<<<< HEAD
obsData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID %in% c("DataSet1", "DataSet2")) %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))

knitr::kable(head(simData), digits = 3, caption = "first rows of example data simData")
knitr::kable(head(obsData), digits = 3, caption = "first rows of example data obsData")

metaData <- attr(exampleDataTimeProfile, "metaData")
knitr::kable(metaData2DataFrame(metaData), caption = "list of meta data:")
```



=======
```{r observed-data}
obsData1 <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet1") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "sd", "caption"))

obsData2 <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet2") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))

obsData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID %in% c("DataSet1", "DataSet2")) %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))


knitr::kable(head(obsData1), digits = 3, caption = "first rows of example data obsData1")
knitr::kable(head(obsData2), digits = 3, caption = "first rows of example data obsData2")

metaData <- attr(exampleDataTimeProfile, "metaData")
knitr::kable(metaData2DataFrame(metaData), caption = "list of meta data:")
```
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

### 2.3.1. Corresponding simulated and observed datasets

<<<<<<< HEAD

In this example we want to display corresponding simulated and observed data with same grouping aesthetics like "color" and "linetype", but with different legends. 
For that they have to be ordered accordingly and the first simulated dataset is the plotted with the same colors as the first observed data set and so on.

(To achive this internally the scales related to the aesthetics defined with variable `groupAesthetics` are restarted after the simulated ggplots layers and the internal counter for color, linetype,.. is started again.)


```{r combine-multiple-simulated-and-observed}
plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping <- aes(
    x = time,
    y = values,
    group = caption
  )
)
```

### 2.3.2 Corresponding simulated and observed datasets with same caption

Note that if observed and simulated data share common groupings, they will be merged in the final legend.

```{r combine-simulated-and-observed-sharing-legend}
# create datasets with common caption
commonSimData <- data.frame(simData) %>%
  dplyr::mutate(caption = gsub("Simulated ", "", caption))

commonObsData <- data.frame(obsData) %>%
  dplyr::mutate(caption = gsub("Observed ", "", caption))

=======
### 2.2.1. Single observed data set

```{r examples-single-observation}
# Define Data Mapping
obsAes <- aes(
  x = time,
  y = values,
  group = caption
)

plotTimeProfile(
  observedData = obsData1,
  metaData = metaData,
  observedMapping = obsAes
)
```



### 2.2.2. Single simulation with confidence interval

For this  data set, the uncertainty/error was defined as standard deviation using the variable `sd`.
If mapped to 'error' this variable will be used to create corresponding `ymin` and `ymax` values for the error bars $ymin = values - sd$, $ymax = values + sd$. 
Also available is error_relative where a multiplicative error is assumed: $ymin = values*error_relative$, $ymax = values*error\_relative$

**Tips**

- For data that include errorbars only partially, `NA` values cannot be used and should be replaced by `0` for error variable or `y` for `ymin` and `ymax` variables. 
- For logarithmic plots, negative `ymin` values of errorbars should be replaced by `y` beforehand in order to still see the `ymax` range of the errorbars.

if the mapping `error` is used and the variable `yscale` is set to 'log', all 'ymin' values < 0 are set to 'y'
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

plotTimeProfile(
  data = commonSimData,
  observedData = commonObsData,
  metaData = metaData,
  mapping <- aes(
    x = time,
    y = values,
    group = caption
  )
)
```


<<<<<<< HEAD
### 2.3.3. Independent simulated and observed datasets

If simulated and observed data sets are not related, set variable `groupAesthetics` to empty vector.
It is now necessary to explicitly map all properties as color, shape and caption

```{r independent-simulated-and-observed-datasets}
# Define Data Mappings
mapping <- aes(
  x = time,
  y = values,
  color = caption,
  fill = caption,
  shape = caption,
  linetype = caption
)

plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = mapping,
  groupAesthetics = c()
)
```


### 2.3.4. Multiple simulations and observed data sets without legends

To map group with an empty variable `groupAesthetics` leads to a plot without legends:

```{r examples-multiple-simulations-no-caption}
mapping <- aes(
  x = time,
  y = values,
  group = caption,
)

plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = mapping,
  groupAesthetics = c()
)
```



### 2.3.5. Multiple simulations and observed data sets with their confidence intervals({#s_multiplePlots})

```{r combine-multiple-simulated-and-observed-with-ci}
# Define Data Mappings
mapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption
)


plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = mapping,
=======
```{r examples-single-observation-with-confidence-interval-as-error}
# Define Data Mapping
obsDataMapping1 <- aes(
  x = time,
  y = values,
  error = sd,
  group = caption
)

plotTimeProfile(
  observedData = obsData1,
  metaData = metaData,
  mapping = obsDataMapping1
)
```


For a second dataset, the uncertainty/error was defined as `"minValues"` and `"maxValues"`, they are input directly as `ymin` and `ymax` and will be taken as is.

```{r examples-single-observation-with-confidence-interval-as-ymin-ymax}
# Define Data Mapping
obsDataMapping2 <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption
)

plotTimeProfile(
  observedData = obsData2,
  metaData = metaData,
  mapping = obsDataMapping2
)
```

### 2.2.3. Multiple observed data sets

```{r examples-multiple-observations}
# Define Data Mapping
obsDataMapping <- aes(
  x = time,
  y = values,
  group = caption
)

plotTimeProfile(
  observedData = obsData,
  metaData = metaData,
  mapping = obsDataMapping
)
```

### 2.2.4. Multiple observed data sets with confidence interval

In this example, the observed data were not using the same way of displaying errors.
Before plotting, the data needs to be merged in a consistent way to let the mapping know which variable(s) to use.

Below, `ymin` and `ymax` were consequently calculated for `obsData1` before merging.

```{r examples-multiple-observations-with-confidence-interval}
# Use common variable before use rbind.data.frame
obsData1$minValues <- obsData1$values - obsData1$sd
obsData1$maxValues <- obsData1$values + obsData1$sd
obsData2$sd <- 0

# Define Data Mapping
obsDataMapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption
)

plotTimeProfile(
  observedData = rbind.data.frame(
    obsData1,
    obsData2
  ),
  metaData = metaData,
  mapping = obsDataMapping
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette
)
```


<<<<<<< HEAD
### 2.3.6 Observed data with shape as gender
=======
### 2.2.5. Missing Dependent Variable (mdv)
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

The following code flags all values higher than `10` as `mdv`, leading to a plot without any observed data points higher than `10` (removing the first observation)

<<<<<<< HEAD
```{r observedData-usead-as-simulatedData}
obsData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet4") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "caption", "gender"))

knitr::kable(head(obsData), digits = 3, caption = "first rows of example data obsData")
```

In this example observed data is used as simulated and as observed data, to connect the different data points:


```{r observedData-usead-as-simulatedData-plot}
plotTimeProfile(
  data = obsData,
  observedData = obsData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    group = caption,
    shape = gender
  ),
  geomLineAttributes = list(linetype = 1, linewidth = 0.5)
) + theme(legend.position = "right")
```
=======

```{r examples mdv}
# Use common variable before usinf rbind.data.frame
mdvData <- obsData1
mdvData$mdv <- mdvData$values > 10

# Define Data Mapping
obsDataMapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption,
  mdv = mdv
)
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

plotTimeProfile(
  observedData = mdvData,
  metaData = metaData,
  mapping = obsDataMapping
)
```


## 2.3. Plot simulated and observed data


```{r observed-and-simulated-data}
simData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID %in% c("DataSet1", "DataSet2")) %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))

<<<<<<< HEAD
## 2.4 Observed data with LLOQ

```{r lloq-data}
simData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == c("DataSet3")) %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::filter(dimension == "concentration") %>%
  dplyr::select(c("time", "values", "caption"))

obsData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet3") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::filter(dimension == "concentration") %>%
  dplyr::select(c("time", "values", "caption", "lloq", "error_relativ"))

knitr::kable(head(simData), digits = 3, caption = "first rows of example data simData")
knitr::kable(head(obsData), digits = 3, caption = "first rows of example data obsData")

metaData <- attr(exampleDataTimeProfile, "metaData")
knitr::kable(metaData2DataFrame(metaData), caption = "list of meta data:")
```

If you have mapped a lloq, a horizontal line for the lloq are displayed, and all values below lloq are plotted with decreased alpha.
Attention we we did some adjustments to get a nicer plot: 

- yscale was set to log, and the y-axis was cut at 0.01
- The simulated line would be normaly in blue, we switched it to black


```{r example-lloq}
plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    group = caption,
    error_relativ = error_relativ,
    lloq = lloq
  ),
  yscale = "log",
  yscale.args = list(limits = c(0.01, NA)),
  geomLineAttributes = list(color = "black")
)
```

## 2.3. Data with Secondary axis


```{r load-data-with-secondary-axis}
simData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet3") %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::select(c("time", "values", "dimension", "caption"))

=======
obsData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID %in% c("DataSet1", "DataSet2")) %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))

knitr::kable(head(simData), digits = 3, caption = "first rows of example data simData")
knitr::kable(head(obsData), digits = 3, caption = "first rows of example data obsData")

metaData <- attr(exampleDataTimeProfile, "metaData")
knitr::kable(metaData2DataFrame(metaData), caption = "list of meta data:")
```




### 2.3.1. Corresponding simulated and observed datasets


In this example we want to display corresponding simulated and observed data with same grouping aesthetics like "color" and "linetype", but with different legends. 
For that they have to be ordered accordingly and the first simulated dataset is the plotted with the same colors as the first observed data set and so on.

(To achieve this internally the scales related to the aesthetics defined with variable `groupAesthetics` are restarted after the simulated ggplots layers and the internal counter for color, linetype,.. is started again.)
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

obsData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet3") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "dimension", "caption", "lloq", "error_relativ"))

<<<<<<< HEAD
knitr::kable(head(obsData), digits = 3, caption = "first rows of example data obsData")

metaData <- attr(exampleDataTimeProfile, "metaData")
knitr::kable(metaData2DataFrame(metaData), caption = "list of meta data:")
```


In this example a plot is generated with concentration on the left y-axis and Fraction on the right y-axis. As first step the variable `metaData` has to be adjusted.
For the primary y-axis the column "values" is mapped, and metaData provides for this column the dimension "Concentration". For the secondary y-axis a new entry "y2" is added.

```{r secondrayAxis_metaData}
metaData <- list(
  time = list(
    dimension = "Time",
    unit = "h"
  ),
  values = list(
    dimension = "Concentration",
    unit = "mg/l"
  ),
  y2 = list(
    dimension = "Fraction",
    unit = ""
  )
)
```


The mapping `y2axis` has to be logical. In this example it is `(dimension == "fraction")`.
For the primary y-axis (concentration) a log scale is displayed and for the secondary (Fraction) a linear scale. Limits of the secondary-axis are set to [0,1]

```{r secondrayAxis_plot}
mapping <- aes(
  x = time,
  y = values,
  shape = caption,
  error_relativ = error_relativ,
  lloq = lloq,
  y2axis = (dimension == "fraction"),
  color = dimension,
  linetype = dimension,
  shape = group
)

plotTimeProfile(
  data = simData,
  observedData = obsData,
  mapping = mapping,
  metaData = metaData,
  yscale = "log",
  yscale.args = list(limits = c(0.01, NA)),
  y2scale = "linear",
  y2scale.args = list(limits = c(0, 1)),
) +
  theme(axis.title.y.right = element_text(angle = 90))
```

# 3. Plot Configuration

```{r observed_and_simulated_data2}
simData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID %in% c("DataSet1", "DataSet2")) %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))

obsData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID %in% c("DataSet1", "DataSet2")) %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))
```


## 3.1 Example for changing geom attributes

The plot from section [Multiple simulations and observed data sets with their confidence intervals](#s_multiplePlots) was adjusted using geom Attributes:

- `geomLineAttributes = list(linetype = 'solid')` The lines use dfor the simulated data are set in both datasets to solid. Attention the linetype for the errorbars and ribbon edges was not changed.
- `geomErrorbarAttributes = list()` The default settings for geomErrorbar, width = 0, was deleted. The bar caps are now visible
- `geomRibbonAttributes = list(alpha = 0.1)` the shade of the ribbons was decreased by setting the alpha to 0.1, the default values for color = NA, was omitted, so that the edges are now visible
- 'geomPointAttributes = list(size = 7)' The size of the symbols are increased


```{r changing_geom_attributes}
# Define Data Mappings
mapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption
)


plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = mapping,
  geomLineAttributes = list(linetype = "solid"),
  geomErrorbarAttributes = list(),
  geomRibbonAttributes = list(alpha = 0.1),
  geomPointAttributes = list(size = 7)
)
```



## 3.2 Example for changing color scales
=======
```{r combine-multiple-simulated-and-observed}
plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping <- aes(
    x = time,
    y = values,
    group = caption
  )
)
```

### 2.3.2 Corresponding simulated and observed datasets with same caption

Note that if observed and simulated data share common groupings, they will be merged in the final legend.

```{r combine-simulated-and-observed-sharing-legend}
# create datasets with common caption
commonSimData <- data.frame(simData) %>%
  dplyr::mutate(caption = gsub("Simulated ", "", caption))

commonObsData <- data.frame(obsData) %>%
  dplyr::mutate(caption = gsub("Observed ", "", caption))


plotTimeProfile(
  data = commonSimData,
  observedData = commonObsData,
  metaData = metaData,
  mapping <- aes(
    x = time,
    y = values,
    group = caption
  )
)
```


### 2.3.3. Independent simulated and observed datasets

If simulated and observed data sets are not related, set variable `groupAesthetics` to empty vector.
It is now necessary to explicitly map all properties as color, shape and caption

```{r independent-simulated-and-observed-datasets}
# Define Data Mappings
mapping <- aes(
  x = time,
  y = values,
  color = caption,
  shape = caption,
  linetype = caption
)

plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = mapping,
  groupAesthetics = c()
)
```


### 2.3.4. Multiple simulations and observed data sets without legends

To map group with an empty variable `groupAesthetics` leads to a plot without legends:

```{r examples-multiple-simulations-no-caption}
mapping <- aes(
  x = time,
  y = values,
  group = caption,
)

plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = mapping,
  groupAesthetics = c()
)
```

>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

For changing scales which ared adressed by the the grouping, the reset of scales after the simulation layers has to be taken in account. (see [Plot simulated and observed data](#s_groupAesthtics)).
It's a stepwise procedure

<<<<<<< HEAD
- call plotTimeprofile for simulated data only
- set scales for simulated data
- call plotTimeprofile for observed data only with the simulated plot as input  `plotObject`
- set scales for observed data


```{r example_changeing_colorscales}
# Define Data Mappings
mapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption
)


plotObject <- plotTimeProfile(
  data = simData,
  metaData = metaData,
  mapping = mapping,
) +
  scale_color_manual(values = c("Simulated Data 1" = "darkred", "Simulated Data 2" = "darkgreen")) +
  scale_fill_manual(values = c("Simulated Data 1" = "red", "Simulated Data 2" = "green"))
=======
### 2.3.5. Multiple simulations and observed data sets with their confidence intervals

```{r combine-multiple-simulated-and-observed-with-ci}
# Define Data Mappings
mapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption
)


plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = mapping,
)
```
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

plotObject <- plotTimeProfile(
  plotObject = plotObject,
  observedData = obsData,
  mapping = mapping
) +
  scale_color_manual(values = c("Observed Data 1" = "darkred", "Observed Data 2" = "darkgreen"))

<<<<<<< HEAD
plotObject
```
=======
### 2.3.6 Observed data with shape as gender
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette


```{r observedData-usead-as-simulatedData}
obsData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet4") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "caption", "gender"))

<<<<<<< HEAD
=======
knitr::kable(head(obsData), digits = 3, caption = "first rows of example data obsData")
```
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

In this example observed data is used as simulated and as observed data, to connect the different data points:


```{r observedData-usead-as-simulatedData-plot}
plotTimeProfile(
  data = obsData,
  observedData = obsData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    group = caption,
    shape = gender
  ),
  geomLineAttributes = list(linetype = 1, linewidth = 0.5)
) + theme(legend.position = "right")
```

<<<<<<< HEAD
## 3.3 Example for adjusting x and y scale
=======
>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette

In the example we set the scale for the y-axis to log scale.
Per default a time profile plots starts always at 0, here the defaults were overwritten and the breaks were set manually.

<<<<<<< HEAD
```{r example_adjsuting_xy_scale}
mapping <- aes(
  x = time + 24,
  y = values,
  group = caption
)

plotTimeProfile(
  data = simData %>%
    dplyr::filter(values > 0),
  metaData = metaData,
  mapping = mapping,
  yscale = "log",
  xscale.args = list(
    limits = c(24, 48),
    breaks = seq(24, 48, 3)
  )
)
```
=======

>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette


## 2.4 Observed data with LLOQ

<<<<<<< HEAD
# 3.3 Adjust Time unit

The breaks of time axis ar set according to the units provided by the variable `metaData`.

Below we show the same plot with 4 different time units

```{r example_timeUnit,fig.show='hold'}
# Define Data Mapping
mapping <- aes(
  x = time,
  y = values,
  group = caption
)

=======
```{r lloq-data}
simData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == c("DataSet3")) %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::filter(dimension == "concentration") %>%
  dplyr::select(c("time", "values", "caption"))

obsData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet3") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::filter(dimension == "concentration") %>%
  dplyr::select(c("time", "values", "caption", "lloq", "error_relative"))

knitr::kable(head(simData), digits = 3, caption = "first rows of example data simData")
knitr::kable(head(obsData), digits = 3, caption = "first rows of example data obsData")

metaData <- attr(exampleDataTimeProfile, "metaData")
knitr::kable(metaData2DataFrame(metaData), caption = "list of meta data:")
```

If you have mapped a `lloq`, a horizontal line for the `lloq` are displayed, and all values below `lloq` are plotted with decreased alpha.
Attention we we did some adjustments to get a nicer plot: 

- `yscale` was set to log, and the y-axis was cut at 0.01
- The simulated line would be by default in blue, we switched it to black


```{r example-lloq}
plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = aes(
    x = time,
    y = values,
    group = caption,
    error_relative = error_relative,
    lloq = lloq
  ),
  yscale = "log",
  yscale.args = list(limits = c(0.01, NA)),
  geomLineAttributes = list(color = "black")
)
```

## 2.3. Data with Secondary axis


```{r load-data-with-secondary-axis}
simData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet3") %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::select(c("time", "values", "dimension", "caption"))


obsData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID == "DataSet3") %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "dimension", "caption", "lloq", "error_relative"))

knitr::kable(head(obsData), digits = 3, caption = "first rows of example data obsData")

metaData <- attr(exampleDataTimeProfile, "metaData")
knitr::kable(metaData2DataFrame(metaData), caption = "list of meta data:")
```


In this example a plot is generated with concentration on the left y-axis and Fraction on the right y-axis. As first step the variable `metaData` has to be adjusted.
For the primary y-axis the column "values" is mapped, and metaData provides for this column the dimension "Concentration". For the secondary y-axis a new entry "y2" is added.

```{r secondrayAxis_metaData}
metaData <- list(
  time = list(
    dimension = "Time",
    unit = "h"
  ),
  values = list(
    dimension = "Concentration",
    unit = "mg/l"
  ),
  y2 = list(
    dimension = "Fraction",
    unit = ""
  )
)
```


The mapping `y2axis` has to be logical. In this example it is `(dimension == "fraction")`.
For the primary y-axis (concentration) a log scale is displayed and for the secondary (Fraction) a linear scale. Limits of the secondary-axis are set to [0,1]

```{r secondrayAxis_plot}
mapping <- aes(
  x = time,
  y = values,
  shape = caption,
  error_relative = error_relative,
  lloq = lloq,
  y2axis = (dimension == "fraction"),
  color = dimension,
  linetype = dimension,
  shape = caption
)

plotTimeProfile(
  data = simData,
  observedData = obsData,
  mapping = mapping,
  metaData = metaData,
  yscale = "log",
  yscale.args = list(limits = c(0.01, NA)),
  y2scale = "linear",
  y2scale.args = list(limits = c(0, 1)),
) +
  theme(axis.title.y.right = element_text(angle = 90))
```

# 3. Plot Configuration

```{r observed_and_simulated_data2}
simData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID %in% c("DataSet1", "DataSet2")) %>%
  dplyr::filter(Type == "simulated") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))

obsData <- exampleDataTimeProfile %>%
  dplyr::filter(SetID %in% c("DataSet1", "DataSet2")) %>%
  dplyr::filter(Type == "observed") %>%
  dplyr::select(c("time", "values", "maxValues", "minValues", "caption"))
```


## 3.1 Example for changing geom attributes

The plot from section [Multiple simulations and observed data sets with their confidence intervals](#s_multiplePlots) was adjusted using geom Attributes:

- `geomLineAttributes = list(linetype = 'solid')` The lines used for the simulated data are set in both datasets to solid. Attention the line type for the errorbars and ribbon edges was not changed.
- `geomErrorbarAttributes = list()` The default settings for `geomErrorbarAttributes`, width = 0, was deleted. The bar caps are now visible
- `geomRibbonAttributes = list(alpha = 0.1)` the shade of the ribbons was decreased by setting the alpha to 0.1, the default values for color = NA, was omitted, so that the edges are now visible
- `geomPointAttributes = list(size = 7)` The size of the symbols are increased


```{r changing_geom_attributes}
# Define Data Mappings
mapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption
)


plotTimeProfile(
  data = simData,
  observedData = obsData,
  metaData = metaData,
  mapping = mapping,
  geomLineAttributes = list(linetype = "solid"),
  geomErrorbarAttributes = list(),
  geomRibbonAttributes = list(alpha = 0.1),
  geomPointAttributes = list(size = 7)
)
```



## 3.2 Example for changing color scales

For changing scales which are addressed by the the grouping, the reset of scales after the simulation layers has to be taken in account. (see [Plot simulated and observed data](#s_groupAesthtics)).
It's a step wise procedure

- call `plotTimeProfile()` for simulated data only
- set scales for simulated data
- call `plotTimeProfile()` for observed data only with the simulated plot as input  `plotObject`
- set scales for observed data


```{r example_changeing_colorscales}
# Define Data Mappings
mapping <- aes(
  x = time,
  y = values,
  ymin = minValues,
  ymax = maxValues,
  group = caption
)


plotObject <- plotTimeProfile(
  data = simData,
  metaData = metaData,
  mapping = mapping,
) +
  scale_color_manual(values = c("Simulated Data 1" = "darkred", "Simulated Data 2" = "darkgreen")) +
  scale_fill_manual(values = c("Simulated Data 1" = "red", "Simulated Data 2" = "green"))

plotObject <- plotTimeProfile(
  plotObject = plotObject,
  observedData = obsData,
  mapping = mapping
) +
  scale_color_manual(values = c("Observed Data 1" = "darkred", "Observed Data 2" = "darkgreen"))

plotObject
```





## 3.3 Example for adjusting x and y scale

In the example we set the scale for the y-axis to log scale.
Per default a time profile plots starts always at 0, here the defaults were overwritten and the breaks were set manually.

```{r example_adjsuting_xy_scale}
mapping <- aes(
  x = time + 24,
  y = values,
  group = caption
)

plotTimeProfile(
  data = simData %>%
    dplyr::filter(values > 0),
  metaData = metaData,
  mapping = mapping,
  yscale = "log",
  xscale.args = list(
    limits = c(24, 48),
    breaks = seq(24, 48, 3)
  )
)
```



# 3.3 Adjust Time unit

The breaks of time axis are set according to the units provided by the variable `metaData`.

Below we show the same plot with 4 different time units:

```{r example_timeUnit,fig.show='hold'}
# Define Data Mapping
mapping <- aes(
  x = time,
  y = values,
  group = caption
)

>>>>>>> 8bc1d35... finalized first version of plotHistogram with Vignette
plotTimeProfile(
  data = simData,
  metaData = metaData,
  mapping = mapping
) +
  labs(
    title = 'Time unit "h"',
    tag = "A"
  )

metaData$time$unit <- "day(s)"

plotTimeProfile(
  data = simData,
  metaData = metaData,
  mapping = mapping
) +
  labs(
    title = 'Time unit "day(s)"',
    tag = "B"
  )


metaData$time$unit <- "weeks(s)"

plotTimeProfile(
  data = simData,
  metaData = metaData,
  mapping = mapping
) +
  labs(
    title = 'Time unit "week(s)"',
    tag = "C"
  )


metaData$time$unit <- "month(s)"

plotTimeProfile(
  data = simData,
  metaData = metaData,
  mapping = mapping
) +
  labs(
    title = 'Time unit "month(s)"',
    tag = "D"
  )
```


